---
title: "supplement 6"
author: "Steve Simon"
date: "May 20, 2018"
output: html_document
---

This file creates various formulas and graphs needed to illustrate the underlying theory behind the data analyses in survival-lecture-6.Rmd. I am sharing this for those who are curious, but you are not responsible for learning or using the code shown in this supplement.


```{r render, eval=FALSE, echo=FALSE}
# run the commands in this section to store the output
# in the results folder.
library(rmarkdown)
f <- "~/survival-models/src/supplement-6.Rmd"
o <- "~/survival-models/results"
render(f, output_dir=o)
```

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

## The Kaplan-Meier plot and proportional hazards.

```{r equal, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
n <- 300
t1 <- rexp(n, 0.6)
t2 <- rexp(n, 0.6)
censor <- as.numeric(c(t1, t2) < 3)
eq <- data.frame(
  time=pmin(c(t1, t2), 3), 
  gp=rep(1:2, each=n), 
  censor=censor)
eq_surv <- Surv(eq$time, eq$censor)
plot(survfit(eq_surv~eq$gp))

```

```{r diverging, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
n <- 300
t1 <- rexp(n, 0.6)
t2 <- rexp(n, 1.2)
censor <- as.numeric(c(t1, t2) < 3)
div03 <- data.frame(
  time=pmin(c(t1, t2), 3), 
  gp=rep(1:2, each=n), 
  censor=censor)
div03_surv <- Surv(div03$time, div03$censor)
plot(survfit(div03_surv~div03$gp), xlim=c(0,10))
```

```{r diverging-10-years, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
n <- 300
t1 <- rexp(n, 0.6)
t2 <- rexp(n, 1.2)
t1[1] <- 9.9
t2[1] <- 9.9
censor <- as.numeric(c(t1, t2) < 10)
div10 <- data.frame(
  time=pmin(c(t1, t2), 10), 
  gp=rep(1:2, each=n), 
  censor=censor)
div10_surv <- Surv(div10$time, div10$censor)
plot(survfit(div10_surv~div10$gp), xlim=c(0,10))
```

A key and very important deviation from proportional hazards is when one group has increased hazard early and a second group has increased hazard late. The Kaplan-Meier curves for the two groups will show this if they cross.

```{r crossing, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
n <- 300
t1 <- rexp(n, 0.6)
t2 <- rgamma(n, 0.2, 0.02)
censor <- as.numeric(c(t1, t2) < 3)
cross <- data.frame(
  time=pmin(c(t1, t2), 3), 
  gp=rep(1:2, each=n), 
  censor=censor)
cross_surv <- Surv(cross$time, cross$censor)
plot(survfit(cross_surv~cross$gp))
```

## Complementary log-log plot

If the proportional hazards assumption holds then the survival curves for two different groups should be related by

$S_2(t) = S_1(t)^{hr}$

Take the logarithm of both sides to get

$log(S_2(t)) = log(S_1(t)) hr$

We'd like to take a second log here, but since $S_2$ and $S_1$ are always between 0 and 1, their logarithms would be negative. You have to flip this to a positive value and then take a second logarithm.

$log(-log(S_2(t))) = log(-log(S_1(t))) + log(hr)$

If this transformation, the complementary log-log transformation, produces two curves that are separated by a constant for all values of t, then you have evidence to support the proportional hazards assumption.

```{r log-log-nice, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
div03_km <- survfit(div03_surv~div03$gp)
g1 <- 1:div03_km$strata[1]
g2 <- (div03_km$strata[1]+1):sum(div03_km$strata)
plot(
  div03_km$time, 
  log(-log(div03_km$surv)), 
  type="n")
lines(
  div03_km$time[g2], 
  log(-log(div03_km$surv[g2])), 
  type="s", lty="dotted")
lines(
  div03_km$time[g1], 
  log(-log(div03_km$surv[g1])), 
  type="s")
```

```{r log-log-nasty, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
cross_km <- survfit(cross_surv~cross$gp)
g1 <- 1:cross_km$strata[1]
g2 <- (cross_km$strata[1]+1):sum(cross_km$strata)
plot(
  cross_km$time, 
  log(-log(cross_km$surv)), 
  type="n")
lines(
  cross_km$time[g2], 
  log(-log(cross_km$surv[g2])), 
  type="s", lty="dotted")
lines(
  cross_km$time[g1], 
  log(-log(cross_km$surv[g1])), 
  type="s")
```

## Review the likelihood ratio test

The partial likelihood is

$l_p = \prod_i \frac{e^{X_{(i)} \beta}}{\sum_{j \in R_{(i)}} e^{X_{j}\beta}}$

The log partial likelihood is

$L_p=\sum_i \big( X_{(i)} \beta - log(\sum_{j \in R_{(i)}} e^{X_j \beta}) \big)$

The derivative of the log partial likelihood is

$\frac{\partial L_p}{\partial \beta}= \sum\limits_{i=1}^m \big( X_{(i)} - \bar{X}_i(\beta) \big)$

where $\bar{X}_i(\beta)$ is a weighted average of all the X's remaining in the risk set and with weights equal to

$w_{ij}=\frac{e^{X_j \beta}}{\sum_{l \in R_i} e^{X_l \beta}}$.

A positive derivative implies that we could maximize the log partial likelihood by increasing from the current value of $\beta$ and a negative derivative implies the opposite.

## Schoenfeld residuals.

The slope of the likelihood function is zero at its maximum, which implies that 

$\sum\limits_{i=1}^m \big( X_{(i)} - \bar{X}_i(\hat{\beta}) \big) = 0$.

The individual terms in this sum,

$X_{(i)} - \bar{X}_i(\hat{\beta})$,

are called the Schoenfeld residuals. 

Normally, the Schoenfeld residuals are standardized. If you see a time trend with respect to the Schoenfeld residuals, then this is evidence of a violation of the proportional hazards assumption.

If the Schoenfeld residuals show a positive time trend (negative early and positive late), that means that you have a hazard ratio that is a bit too large early and a bit too small late.

## Stratified analysis

If one of the variables in your model is a nuisance variable, but it could possibly be the cause of the violation of the proportional hazards assumption, you can fit fix this by fitting a separate baseline hazard for each level of your nuisance variable.

```{r stratified-analysis}
fn <- "~/survival-models/bin/whas500.RData"
load(fn)
year_km <- survfit(
  Surv(whas500$time_yrs, whas500$fstat=="Dead")~whas500$year)
plot(year_km)
```

## Time-varying covariates

Page 221 of Hosmer, Lemeshow, and May describes the grace data set, which you can use for time-varying covariates.

```{r read-grace}
fn <- "~/survival-models/data/wiley/GRACE1000.dat" 
grace <- read.table(fn, header = FALSE) 
names(grace) <- c(
  "id",
  "days",
  "death",
  "revasc",
  "revascdays",
  "los",
  "age",
  "sysbp",
  "stchange")
head(grace)
```

## Review the likelihood ratio test

Consider a small data set with the following survival times:

1 M   6 1

2 F  14 1

3 F  44 1

4 M  45 0

5 F  89 0

6 M  98 1

7 F  99 1

8 F 104 1

9 M 114 1

the partial likelihood ratio is

$l_p = \prod_i \frac{\psi_{(i)}}{\sum_{j \in R_{(i)}} \psi_j}$

where the parentheses around the subscript implicitly excludes those patients who were censored,

$\psi_{(i)} = e^{X_{(i)} \beta}$

and $R_{(i)}$ is the set of  all patients at risk at time $t_{(i)}$.

For a simple case with a binary predictor like gender, you would have $\psi_{(i)}$ equalling a constant $\psi$ for females and a constant 1 for males.

It helps to keep tabs on the running number of men and women at risk at each time point.

1 M   6 1 4/5

2 F  14 1 3/5

3 F  44 1 3/4

4 M  45 0 3/3

5 F  89 0 2/3

6 M  98 1 2/2

7 F  99 1 1/2

8 F 104 1 1/1

9 M 114 1 1/0

At t=6, a male died, making the numerator 1. The denominator is 4+5$\psi$ because there are 4 males and 5 females in the risk set. The fraction is

$\frac{1}{4+5\psi}$

At t=14, a female died, making the numerator $\psi$ and the denominator is 3+5$\psi$ because there are 3 males and 5 females in the risk set. The second fraction in the product is

$\frac{\psi}{3+5\psi}$

At t=44, another female died, making the numerator $\psi$ again, but the denominator is 3+4$\psi$ because there are 3 males and 4 females in the risk set. This fraction is

$\frac{\psi}{3+4\psi}$

You skip the censored values of t=45 and t=89. The remaining fractions are

$\frac{1}{2+2\psi}$

$\frac{\psi}{1+2\psi}$

$\frac{\psi}{1+\psi}$

$\frac{1}{1}$

The entire product works out to be

$\frac{\psi^4}{(4+5\psi)(3+5\psi)(3+4\psi)(2+2\psi)(1+2\psi)(1+\psi)}$

Let's plot this for various values of $\psi=e^\beta$.

```{r plot-lr}
lp <- function(beta) {
  psi <- exp(beta)
  f <- psi^4 / 
    ((4+5*psi)*
     (3+5*psi)*
     (3+4*psi)*
     (2+2*psi)*
     (1+2*psi)*
     (1+  psi))
  return(log(f))
}
beta <- seq(-2, 2, length=100)
plot(beta, lp(beta), type="l")
beta[which.max(lp(beta))]
```

The second derivative of the log partial likelihood is

$\frac{\partial L_p}{\partial \beta}= -\sum_i \sum_{j \in R_i}{} w_{ij}\big( X_j - \bar{X}_{w{(i)}} \big)^2$

which is a weighted variance.

The information matrix is the negative of the second derivative and the inverse of the information matrix gives you the variances and covariances of the maximum likelihood estimates.

## Also for discussion

* Stratified analysis. Stratify whas500 by year=1, 2, 3. This would use the actg320 data. Compare output to Table 7.1.

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/supplement-6.RData")
```