---
title: "Model fitting and diagnostics for the Cox model"
author: "Steve Simon"
date: "April 8, 2018"
output: 
  powerpoint_presentation
---

## Abstract

Lecture 4. Model fitting and diagnostics for the Cox model. In this lecture, you will work with more complex forms of the Cox model with multiple predictor variables. You'll include covariates in the Cox model to produce risk adjusted survival curves. You will also assess the underlying assumptions of the Cox model, particularly the assumption of proportional hazards.

```{r load-libraries, echo=FALSE}
options(width=50)
library(knitr)
knitr::opts_chunk$set(echo=FALSE)
library(broom)
library(dplyr, warn.conflicts=FALSE)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr, warn.conflicts=FALSE, quietly=TRUE)
```

## Advantages of a multivariate model

1. Your predictions are better with two (or more) independent variables.

2. You can use covariates to make risk adjustments.

3. You can explore interactions among variables.


## Review the whas500 data set

```{r read}
fn <- "../../data/whas500.RData" 
load(fn)
whas500$time_yrs <- whas500$lenfol/365.25
head(whas500, n=2)
```

## Model fitting strategies

1. Fit univariate models first.

2. Add variables one at a time or in very small batches.

3. Look at interactions and nonlinearities last.

## Univariate model for age

```{r age-model}
display_ci <- function(tidy_object) {
  tidy_object %>%
    mutate(conf.int=paste0(
      round(exp(conf.low), 2),
      "-",
      round(exp(conf.high), 2)
    ))
}
display_pv <- function(tidy_object) {
  tidy_object %>%
    mutate(p.value=round(p.value, 3)) %>%
    mutate(p.value=pmax(p.value, 0.001))    
}
clean_table <- function(cox_object) {
  cox_object %>%
    tidy %>%
    mutate(hr=round(exp(estimate), 2)) %>%
    display_ci %>%
    display_pv %>%
    select(
      term,
      hr,
      p.value,
      conf.int)
}
cox_age <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age,
    data=whas500)
clean_table(cox_age)
```

## Univariate model for gender

```{r gender-model}
cox_gender <- 
  coxph(Surv(time_yrs, fstat=="Dead")~gender,
    data=whas500)
clean_table(cox_gender)
```

## Estimated survival by gender

```{r gender-plot, fig.width=4.5, fig.height=2.5}
n <- data.frame(gender=factor(1:2, labels=c("Male", "Female")))
cox_predict <- survfit(cox_gender, newdata=n)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  cox_predict$time,
  cox_predict$surv[, 2], 
  type="s", ylim=c(0,1))
lines(
  cox_predict$time, 
  cox_predict$surv[, 1], 
  type="s", lty="dotted")
five_year <- which.min(abs(cox_predict$time-5))
y <- cox_predict$surv[five_year, ]+0.05
text(5, y, c("M", "F"))
```

## Model with age and gender

```{r age-and-gender}
cox_mv <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age+gender,
    data=whas500)
clean_table(cox_mv)
```

## What is happening here?

The average age across all subjects is `r round(mean(whas500$age), 1)`, but the averages by gender are quite different. For males, the average age is `r round(mean(whas500$age[whas500$gender=="Male"]), 1)`, but for females, the average age is `r round(mean(whas500$age[whas500$gender=="Female"]), 1)`.

## Boxplots of age by gender

```{r boxplots, fig.width=4.5, fig.height=2.5}
overall_mean <- mean(whas500$age)
gender_means <- tapply(whas500$age, whas500$gender, mean)
age_diff <- gender_means[2]-gender_means[1]
ggplot(whas500, aes(x=gender, y=age)) +
  geom_boxplot()
coef_mv <- tidy(cox_mv)
hr_age <- exp(coef_mv[coef_mv$term=="age", "estimate"])
extrapolated_hr <- hr_age^age_diff
```

## Adjusting for covariate imbalance

There is a `r round(age_diff, 1)` year difference between the average ages of men and women. The hazard ratio for age, `r round(hr_age, 3)`, can get extrapolated to a `r round(age_diff, 1)` year difference by exponentiating. That is `r round(hr_age, 3)`^`r round(age_diff, 1)` = `r round(extrapolated_hr, 2)` which is actually larger than the hazard ratio that we saw for the unadjusted model with just gender.

## 66.6 year male versus 74.7 year female

```{r unadjusted-graph, fig.width=4.5, fig.height=2.5}
n_unadjusted <- 
  data.frame(
    age=gender_means, 
    gender=factor(1:2, labels=c("Male", "Female")))
predict_unadjusted <- survfit(cox_mv, newdata=n_unadjusted)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_unadjusted$time,
  predict_unadjusted$surv[, 2], 
  type="s", ylim=c(0,1))
lines(
  predict_unadjusted$time, 
  predict_unadjusted$surv[, 1], 
  type="s", lty="dotted")
five_year <- which.min(abs(predict_unadjusted$time-5))
y <- predict_unadjusted$surv[five_year, ]+0.05
text(5, y, c("M", "F"))
```

## 69.8 year male versus 69.8 year female

```{r adjusted-graph, fig.width=4.5, fig.height=2.5}
n_adjusted <- 
  data.frame(
    age=overall_mean,
    gender=factor(1:2, labels=c("Male", "Female")))
predict_adjusted <- survfit(cox_mv, newdata=n_adjusted)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_adjusted$time,
  predict_adjusted$surv[, 2], 
  type="s", ylim=c(0,1))
lines(
  predict_adjusted$time, 
  predict_adjusted$surv[, 1], 
  type="s", lty="dotted")
five_year <- which.min(abs(predict_adjusted$time-5))
y <- predict_adjusted$surv[five_year, ]+0.05
text(5, y, c("M", "F"))
```

## Univariate analysis of bmi

```{r compare-age-and-bmi, fig.width=4.5, fig.height=2.5}
whas500$i_female <- as.numeric(whas500$gender=="Female")
cox_bmi <- 
  coxph(Surv(time_yrs, fstat=="Dead")~bmi,
    data=whas500)
clean_table(cox_bmi)
```

## Unadusted survival curves for different bmi values

```{r bmi-plots, fig.width=4.5, fig.height=2.5}
n_bmi <- data.frame(bmi=c(20, 30, 40))
predict_bmi <- survfit(cox_bmi, newdata=n_bmi)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_bmi$time,
  predict_bmi$surv[, 1], 
  type="s", ylim=c(0,1))
lines(
  predict_bmi$time, 
  predict_bmi$surv[, 2], 
  type="s", lty="dotted")
lines(
  predict_bmi$time, 
  predict_bmi$surv[, 3], 
  type="s", lty="dashed")
five_year <- which.min(abs(predict_bmi$time-5))
y <- predict_bmi$surv[five_year, ]+0.05
text(5, y, c(20, 30, 40))
ggplot(data=whas500) +
  geom_boxplot(aes(x=gender, y=bmi))
ggplot(data=whas500) +
  geom_point(aes(x=age, y=bmi))
```

## Adjusting bmi for age, gender

```{r bmi-adjusted}
cox_bmi_adj <- 
  coxph(Surv(time_yrs, fstat=="Dead")~bmi+age+i_female,
    data=whas500)
clean_table(cox_bmi_adj)
```

## Adjusted bmi survival plots

```{r bmi-plots-adjusted, fig.width=4.5, fig.height=2.5}
n_bmi <- data.frame(age=overall_mean, i_female=1, bmi=c(20, 30, 40))
predict_bmi <- survfit(cox_bmi_adj, newdata=n_bmi)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_bmi$time,
  predict_bmi$surv[, 1], 
  type="s", ylim=c(0,1))
lines(
  predict_bmi$time, 
  predict_bmi$surv[, 2], 
  type="s", lty="dotted")
lines(
  predict_bmi$time, 
  predict_bmi$surv[, 3], 
  type="s", lty="dashed")
five_year <- which.min(abs(predict_bmi$time-5))
y <- predict_bmi$surv[five_year, ]+0.05
text(5, y, c(20, 30, 40))
```

## An interaction model; the raw interaction is hard to interpret

```{r int}
cox_int1 <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age*i_female,
    data=whas500)
clean_table(cox_int1)
```

## Interaction using centered values is easier to interpret

```{r int-centered}
whas500$age_c <- whas500$age-overall_mean
cox_int2 <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age_c*i_female,
    data=whas500)
clean_table(cox_int2)
```

## Gender hazard ratio by age

```{r int-plot, fig.width=4.5, fig.height=2.5}
coef_int2 <- tidy(cox_int2)$estimate
hr <- exp(coef_int2[2]+coef_int2[3]*((40:90)-overall_mean))
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(40:90, hr, type="l")
abline(h=1, lty="dotted")
```

## You can use a sequence of Wald tests to compare different models

```{r wald}
cox_v1 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    gender,
      data=whas500)

cox_v2 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    age+gender,
      data=whas500)                

cox_v3 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    bmi+age+gender,
      data=whas500)

cat("Model 1")
clean_table(cox_v1)
cat("\nModel 2")
clean_table(cox_v2)
cat("\nModel 3")
clean_table(cox_v3)
```

## Comparing using likelihoods

You use the log partial likelihood and/or the AIC (Akaike Information Criteria) to compare models of different complexity.

AIC = -2 log Likelihood + 2 k.

AIC = -2 log Likelihood + log(n) k.

## AIC comparisons

```{r aic}
cox_v1                           %>%
  glance                         %>%
  mutate(lab="gender only")      -> aic1

cox_v2                           %>%
  glance                         %>%
  mutate(lab="gender, age")      -> aic2

cox_v3                           %>%
  glance                         %>%
  mutate(lab="gender, age, bmi") -> aic3

aic1                             %>%
  bind_rows(aic2)                %>%
  bind_rows(aic3)                %>%
  select(lab, logLik, AIC, BIC)  -> aic_all
aic_all
```

## Martingale residuals

There are several residuals available for Cox regression. The Martingale residual is defined as

$M(t_i) = \delta_i - H_0(t_i) e^{X\beta}$

where $\delta_i$ = 0 if censored, 1 if dead.

## Residual plot for age

```{r residuals-age, fig.width=4.5, fig.height=2.5}
whas500$r <- residuals(cox_v3)
par(mar=c(2.6, 2.6, 0.6, 0.6))
res_plot <- ggplot(whas500, aes(age, r)) +
  geom_point()
suppressMessages(print(res_plot))
```

## Residual plot for age with smoothing line

```{r residuals-age-smoothed, fig.width=4.5, fig.height=2.5}
res_plot <- ggplot(data=whas500, aes(age, r)) +
  geom_point() +
  geom_smooth(se=TRUE)
suppressMessages(print(res_plot))
```

## Residual plots for hr

```{r residuals-hr}
res_plot <- ggplot(data=whas500, aes(hr, r)) +
  geom_point() +
  geom_smooth(se=TRUE)
suppressMessages(print(res_plot))
res_plot <- ggplot(data=whas500, aes(sysbp, r)) +
  geom_point() +
  geom_smooth(se=TRUE)
suppressMessages(print(res_plot))
res_plot <- ggplot(data=whas500, aes(diasbp, r)) +
  geom_point() +
  geom_smooth(se=TRUE)
suppressMessages(print(res_plot))
res_plot <- ggplot(data=whas500, aes(cvd, r)) +
  geom_boxplot()
suppressMessages(print(res_plot))
res_plot <- ggplot(data=whas500, aes(afb, r)) +
  geom_boxplot()
suppressMessages(print(res_plot))
res_plot <- ggplot(data=whas500, aes(sho, r)) +
  geom_boxplot()
suppressMessages(print(res_plot))
res_plot <- ggplot(data=whas500, aes(chf, r)) +
  geom_boxplot()
suppressMessages(print(res_plot))
res_plot <- ggplot(data=whas500, aes(av3, r)) +
  geom_boxplot()
suppressMessages(print(res_plot))
res_plot <- ggplot(data=whas500, aes(miord, r)) +
  geom_boxplot()
suppressMessages(print(res_plot))
res_plot <- ggplot(data=whas500, aes(mitype, r)) +
  geom_boxplot()
suppressMessages(print(res_plot))
```

## Using splines to model non-linearities

```{r spline, fig.width=4.5, fig.height=2.5}
cox_pspline3 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    bmi+pspline(age, df=3)+gender,
      data=whas500)
terms_pspline3 <- predict(cox_pspline3, type="terms")
par(mar=c(2.6, 2.6, 0.6, 0.6))
o <- order(whas500$age)
plot(whas500$age[o], terms_pspline3[o ,2], type="l")
```

## An 8 df spline (overfitting?)

```{r overfitting, fig.width=4.5, fig.height=2.5}
cox_pspline8 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    bmi+pspline(age, df=8)+gender,
      data=whas500)
terms_pspline8 <- predict(cox_pspline8, type="terms")

par(mar=c(2.6, 2.6, 0.6, 0.6))
plot(whas500$age[o], terms_pspline8[o ,2], type="l")
```

## Comparing linear model to two splines

```{r comparing-splines}
glance(cox_bmi_adj) %>%
  bind_rows(glance(cox_pspline3)) %>%
  bind_rows(glance(cox_pspline8)) %>%
  mutate(lab=c(
    "linear (df=1)",
    "spline (df=3)",
    "spline (df=8)"))             %>%
  select(lab, logLik, AIC, BIC)   -> compare_splines
compare_splines
```


## Next time - testing proportional hazards

1. Patterns in Kaplan-Meier curves

2. Complementary log-log plot

3. Schoenfeld Residuals

4. Fit time varying covariates

## What have you learned today?

1. The Cox regression model allows for multiple independent variables and interactions.

2. The predicted survival curve estimated at a common covariate mean produces a risk-adjusted comparison.

3. A positive martingale residual implies a death earlier than expected.

```{r save-everything}
save.image("../../data/class4.RData")
```
