---
title: "homework answers 5"
author: "Steve Simon"
date: "May 27, 2018"
output: html_document
---

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

Open the WHAS500 data set in the software program of 
your choice.

```{r open}
fn <- "~/survival-models/bin/whas500.RData"
load(fn)
head(whas500)
```

Calculate and graph on the same graph an overall Kaplan-Meier survival curve, a survival curve for an exponential model with no independent variables, and a survival curve for a Weibull model with no independent variables.

```{r graph}
whas500_surv <- Surv(whas500$time_yrs, whas500$fstat=="Dead")
km_overall <- survfit(whas500_surv~1)

exp_null_model <- survreg(whas500_surv~1, dist="exponential")
pct <- (1:999)/1000
exp_null_model %>%
  predict(
    newdata=data.frame(one=1),
    type="quantile",
    p=pct)                     %>%
  data.frame                   %>%
  set_names("time")            %>%
  mutate(estimate=rev(pct))    %>%
  mutate(grp="Exponential")        %>%
  filter(time<=10)             -> exp_survival
weibull_null_model <- survreg(whas500_surv~1, dist="weibull")
weibull_null_model %>%
  predict(
    newdata=data.frame(one=1),
    type="quantile",
    p=pct)                     %>%
  data.frame                   %>%
  set_names("time")            %>%
  mutate(estimate=rev(pct))    %>%
  mutate(grp="Weibull")        %>%
  filter(time<=10)             -> weibull_survival

km_overall                     %>%
  tidy                         %>%
  mutate(grp="Kaplan-Meier")   %>%
  bind_rows(exp_survival)      %>%
  bind_rows(weibull_survival)  -> combined_survivals

combined_survivals             %>%
  ggplot(aes(time, estimate))   +
    expand_limits(y=0)          +
    geom_step(aes(color=grp))

```

Does there appear to be a difference between the exponential model and the Weibull model? Does either model match the Kaplan-Meier curve?

Answer: The exponential model is a poor fit. It overestimates survival early and underestimates it late. The Weibull model is not perfect, but it matches much more closely.

Calculate a Weibull regression model with bmi, age, gender, and a age by gender interaction. Estimate the impact of a 5 unit change in BMI on the survival percentiles, holding all of the other variables constant.

```{r fit-weibull}
weibull_full_model <- survreg(
  whas500_surv~bmi + age*gender, 
  data=whas500,
  dist="weibull")
summary(weibull_full_model)
weibull_full_model          %>%
  tidy                      %>%
  filter(term=="bmi")       %>%
  select(estimate)          %>%
  mutate(hr=exp(estimate))  %>%
  mutate(hr5=hr^5)
```

Calculate a confidence interval for the scale parameter of the Weibull model.

```{r scale-ci}
weibull_full_model                           %>%
  tidy                                       %>%
  filter(term=="Log(scale)")                 %>%
  select(estimate, std.error)                %>%
  mutate(scale=exp(estimate))                %>%
  mutate(lo=exp(estimate-1.96*std.error))    %>%
  mutate(hi=exp(estimate+1.96*std.error))
```

Interpret this interval.

Answer: Since this interval excludes the value of 1, you can conclude that the exponential model is not a good fit for the data.

1. Open the file, diabetes.csv.

```{r load}
library(timereg)
data(diabetes)
head(diabetes)
diabetes_surv <- Surv(diabetes$time, diabetes$status)
```

I cheated here and took the file directly from the timereg library.

a. Calculate and graph a Kaplan-Meier curve comparing treated to untreated eyes, ignoring for now the correlations inherent in this data set. Does it appear as if these survival curves differ? If so, do they appear to violate the assumption of proportional hazards?

```{r km, fig.width=4.5, fig.height=2.5}
survfit(diabetes_surv~treat, data=diabetes) %>%
  tidy                                      %>%
  ggplot(aes(time, estimate, color=strata))  +
  expand_limits(y=0)                         +
  geom_step()
```

The curves doe seem to differ, but do not appear to violate the assumption of proportional hazards.

b. Calculate and interpret a Cox regression model using treat as an independent variable and id as a cluster effect.

```{r cluster}
coxph(
  diabetes_surv~treat+agedx+cluster(id),
  data=diabetes)
```

There is a statistically significant treatment effect, even after allowing for the correlation within subjects. The robust standard error appears to be slightly smaller, which means that accounting for the pairing has improved your precision.

c. Calculate and interpret a Cox regression model using treat as an independent variable and id as a frailty effect.

```{r frailty}
coxph(
  diabetes_surv~treat+agedx+frailty(id),
  data=diabetes)
```

Normally, you don't fit both a cluster and a frailty model. It is better to pick one prior to data collection and stick with it. Nevertheless, the conclusions remain largely the same.



Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/homework-answers-5.RData")
```