---
title: "homework answers 6"
author: "Steve Simon"
date: "May 27, 2018"
output: html_document
---

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

Open the WHAS500 data set in the software program of your choice.

```{r read}
fn <- "~/survival-models/bin/whas500.RData"
load(fn)
```

Calculate and graph on the same graph a Kaplan-Meier curve for the three cohorts associated with year. Does it appear as if these survival curves differ? If so, do they appear to violate the assumption of proportional hazards?

```{r km-plots, fig.width=4.5, fig.height=2.5}
whas500_surv <- Surv(whas500$time_yrs, whas500$fstat=="Dead")
km_by_year <- survfit(whas500_surv~whas500$year)
km_by_year                                 %>%
  tidy                                     %>%
  ggplot(aes(time, estimate, color=strata)) +
  geom_step()
```

Calculate and interpret the complementary log-log plots.

```{r log-log, fig.width=4.5, fig.height=2.5}
km_by_year                            %>%
  tidy                                %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()
```

Calculate, plot, and interpret the Schoenfeld residuals from a Cox regression model with year as the only independent variable. 

```{r schoenfeld, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 4.1, 0.6, 0.6))
cox_mv <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ year, data=whas500)
cox_schoenfeld <- cox.zph(cox_mv)
print(cox_schoenfeld)
plot(cox_schoenfeld)
```

Fit a Cox regression model with gender as an independent variable and include year as a strata. Create estimated survival plots for each strata comparing males to females.

```{r stratified model, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 4.1, 0.6, 0.6))
cox_stratified <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ gender + strata(year), data=whas500)
n <- data.frame(
  gender=c("Male", "Female", "Male", "Female", "Male", "Female"),
  year= c("1997", "1997", "1999", "1999", "2001", "2001"))

cox_predict <- survfit(cox_stratified, newdata=n)

cox_predict                                            %>%
  tidy                                                 %>%
  filter(strata=="1" | strata=="2")                    %>%
  mutate(gender=ifelse(strata=="1", "Male", "Female")) %>%
  ggplot(aes(time, estimate, color=gender))             +
    expand_limits(y=0)                                  +
    geom_step()

cox_predict                                            %>%
  tidy                                                 %>%
  filter(strata=="3" | strata=="4")                    %>%
  mutate(gender=ifelse(strata=="3", "Male", "Female")) %>%
  ggplot(aes(time, estimate, color=gender))             +
    expand_limits(y=0)                                  +
    geom_step()

cox_predict                                            %>%
  tidy                                                 %>%
  filter(strata=="5" | strata=="6")                    %>%
  mutate(gender=ifelse(strata=="5", "Male", "Female")) %>%
  ggplot(aes(time, estimate, color=gender))             +
    expand_limits(y=0)                                  +
    geom_step()
```

1. Open the file, leader.txt

Calculate a competing risk model comparing Latin America to the other three regions. What are the major differences?

```{r read}
fn <- "~/survival-models/bin/leader.RData"
load(fn)
leader <- leader[is.finite(leader$age), ]
surv_object <- 
  Surv(
    leader$years, 
    leader$lost,
    type="mstate")
```

```{r subsets-2, fig.width=4.5, fig.height=2.5}
sb3 <- leader$region=="latin america"
km_sb3 <- survfit(surv_object~1, subset=sb3)
ggplot(
  data=tidy(km_sb3), 
  aes(time, estimate, fill=state)) +
  expand_limits(y=1)               +
  geom_col(width=1)
sb4 <- leader$region!="latin america"
km_sb4 <- survfit(surv_object~1, subset=sb4)
ggplot(
  data=tidy(km_sb4), 
  aes(time, estimate, fill=state)) +
  expand_limits(y=1)               +
  geom_col(width=1)
```

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/homework-answers-6.RData")
```