---
title: "homework1"
author: "Steve Simon"
date: "April 23, 2018"
output: html_document
---

Preliminaries.

```{r prelims}
library(survival)
```

Read in the WHAS500 data set.

```{r read}
fn <- "../../data/wiley/whas500.dat"
whas500 <- read.table(fn, header=FALSE, as.is=TRUE)
names(whas500) <- c(
  "id",
  "age", 
  "gender",
  "hr",
  "sysbp",
  "diasbp",
  "bmi",
  "cvd",
  "afb",
  "sho",
  "chf",
  "av3",
  "miord",
  "mitype",
  "year",
  "admitdate",
  "disdate",
  "fdate",
  "los",
  "dstat",
  "lenfol",
  "fstat")
whas500$time_yrs <- whas500$lenfol / 365.25
whas500$gender <- 
  factor(whas500$gender, levels=0:1,
    labels=c("Male", "Female"))
whas500$cvd <- 
  factor(whas500$cvd, levels=0:1,
    labels=c("No", "Yes"))
whas500$afb <- 
  factor(whas500$afb, levels=0:1,
    labels=c("No", "Yes"))
whas500$sho <- 
  factor(whas500$sho, levels=0:1,
    labels=c("No", "Yes"))
whas500$cvd <- 
  factor(whas500$fstat, levels=0:1,
    labels=c("No", "Yes"))
whas500$chf <- 
  factor(whas500$chf, levels=0:1,
    labels=c("No", "Yes"))
whas500$av3 <- 
  factor(whas500$av3, levels=0:1,
    labels=c("No", "Yes"))
whas500$miord <- 
  factor(whas500$miord, levels=0:1,
    labels=c("First", "Recurrent"))
whas500$mitype <- 
  factor(whas500$mitype, levels=0:1,
    labels=c("Non Q-wave", "Q-wave"))
whas500$year <- 
  factor(whas500$chf, levels=1:3,
    labels=c("1997", "1999", "2001"))
whas500$dstat <- 
  factor(whas500$dstat, levels=0:1,
    labels=c("Alive", "Dead"))
whas500$fstat <- 
  factor(whas500$fstat, levels=0:1,
    labels=c("Alive", "Dead"))
head(whas500)
fn <- "../../data/whas500.RData"
save(whas500, file=fn)
```

Produce a table of counts for fstat.

```{r counts}
table(whas500$fstat)
```

Draw a Kaplan-Meier plot for overall survival.

```{r overall-km, fig.width=5, fig.height=5}
whas500_surv <- 
  Surv(whas500$lenfol, whas500$fstat=="Dead")
whas500_km <- survfit(whas500_surv~1)
plot(whas500_km, conf.int=FALSE)
```

Estimate the 25th, 50th, and 75th quantiles for overall survival.

```{r quantiles}
quantile(whas500_km)
```

Produce a crosstabulation of fstat and gender. Are you comfortable with the number of deaths in each group?

```{r crosstab}
table(whas500$gender, whas500$fstat)
```

Draw Kaplan-Meier curves for males and females.

```{r km-by-gender}
whas500_by_gender <- survfit(whas500_surv~whas500$gender)
plot(whas500_by_gender)
```

Calculate median survival with confidence intervals
for males and females.

```{r q-by-gender}
quantile(whas500_by_gender)
```

Calculate the log rank test for males versus females. Interpret your result.

```{r logrank-by-gender}
survdiff(whas500_surv~whas500$gender)
```

Interpretation: the p-value is less than 0.05, so you would reject the null hypothesis and conclude that the survival probabilities are different for women versus men.

Produce age groups <60, 60-69, 70-79, and >=80. 
Compute a crosstabulation of this variable with fstat.
Are you comfortable with the number of deaths in each
group?

```{r cutpoints-for-age}
library(broom)
library(ggplot2)
library(magrittr)
library(tidyr)
age_breaks <- c(0, 59, 69, 79, 99)
age_labels <- c("<60", "60-69", "70-79", ">=80")
whas500$age_group <- cut(whas500$age, age_breaks, age_labels)
table(whas500$age_group, whas500$fstat)
```

Draw Kaplan Meier curves for each age group.

```{r km-by-age}
whas500_km_by_age <- survfit(whas500_surv~whas500$age_group)
whas500_km_by_age    %>%
  tidy               %>%
  ggplot(aes(time, estimate, color=strata)) +
    geom_step()
```

Calculate the median survival time with confidence intervals for each age group.

```{r quantile-by-age}
quantile(whas500_km_by_age)
```

Calculate the log rank test for age groups. Interpret your results.

```{r logrank-by-age}
survdiff(whas500_surv~whas500$age_group)
```


(Only for those who are brave) The following are 
times for catheters in infants. A "+" means that the
catheter was removed because it was no longer needed.
Times without a + mean that the catheter was removed
because it failed. Occlusion and infection were the 
two major reasons for failure. Treating failures as
an event and removal because it was no longer needed
as a censored observation, estimate the Kaplan-Meier
survival curve by hand, showing all your intermediate
calculations.

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, 2+, 2+, 2, 2, 
3+, 3, 4+, 4, 5+, 5+, 5+, 5+, 5+, 5, 5, 5, 6, 6, 7, 
10, 10, 12, 12, 13

First, set up a data frame for the ten time points (1-7, 10, 12, 13)

```{r set-up}
km <- data.frame(
  t=c(1:7, 10, 12, 13),
  n=rep(-1, 10),
  d=rep(-1, 10),
  c=rep(-1, 10),
  p=rep(-1, 10),
  s=rep(-1, 10))
```

We will fill in these numbers soon enough. Here's the key:
  t = time
  n = number at risk
  d = number of failures
  c = number censored
  p = conditional probability
  s = survival probability
  
**1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1,** 2+, 2+, 2, 2, 
3+, 3, 4+, 4, 5+, 5+, 5+, 5+, 5+, 5, 5, 5, 6, 6, 7, 
10, 10, 12, 12, 13

There are 34 observations total. At t1=1, there are d1=2 deaths and c1=8 ceonsored values. The survival probability is equal to the conditional probability.

```{r t1}
km$n[1] <- 34
km$d[1] <-  2 
km$c[1] <-  8
km$p[1] <- 1-km$d[1]/km$n[1]
km$s[1] <- km$p[1]
km
```

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, **2+, 2+, 2, 2,** 
3+, 3, 4+, 4, 5+, 5+, 5+, 5+, 5+, 5, 5, 5, 6, 6, 7, 
10, 10, 12, 12, 13

At t2=2, there are n2=24 at risk, and you have d2=2 failures and c2=2 censored observations. The survival probability is equal to the product of the first two conditional probabilities.

```{r t2}
km$n[2] <- 24
km$d[2] <-  2 
km$c[2] <-  2
km$p[2] <- 1-km$d[2]/km$n[2]
km$s[2] <- km$p[1]*km$p[2]
km
```

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, 2+, 2+, 2, 2, 
**3+, 3,** 4+, 4, 5+, 5+, 5+, 5+, 5+, 5, 5, 5, 6, 6, 7, 
10, 10, 12, 12, 13

At t3=3, there are n3=20 at risk, and you have d3=1 failure and c3=1 censored observation. The survival probability is equal to the product of the first three conditional probabilities.

```{r t3}
km$n[3] <- 20
km$d[3] <-  1 
km$c[3] <-  1
km$p[3] <- 1-km$d[3]/km$n[3]
km$s[3] <- km$p[1]*km$p[2]*km$p[3]
km
```

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, 2+, 2+, 2, 2, 
3+, 3, **4+, 4,** 5+, 5+, 5+, 5+, 5+, 5, 5, 5, 6, 6, 7, 
10, 10, 12, 12, 13

At t4=4, there are n4=18 at risk, and you have d4=1 failure and c4=1 censored observations. The survival probability is equal to the product of the first four conditional probabilities.

```{r t4}
km$n[4] <- 18
km$d[4] <-  1 
km$c[4] <-  1
km$p[4] <- 1-km$d[4]/km$n[4]
km$s[4] <- km$p[1]*km$p[2]*km$p[3]*km$p[4]
km
```

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, 2+, 2+, 2, 2, 
3+, 3, 4+, 4, **5+, 5+, 5+, 5+, 5+, 5, 5, 5,** 6, 6, 7, 
10, 10, 12, 12, 13

At t5=5, there are n5=16 at risk, and you have d5=3 failures and c5=5 censored observations. The survival probability is equal to the product of the first five conditional probabilities.

```{r t5}
km$n[5] <- 16
km$d[5] <-  3 
km$c[5] <-  5
km$p[5] <- 1-km$d[5]/km$n[5]
km$s[5] <- km$p[1]*km$p[2]*km$p[3]*km$p[4]*km$p[5]
km
```

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, 2+, 2+, 2, 2, 
3+, 3, 4+, 4, 5+, 5+, 5+, 5+, 5+, 5, 5, 5, **6, 6,** 7, 
10, 10, 12, 12, 13

At t6=6, there are n6=8 at risk, and you have d6=2 failures and c6=0 censored observations. The survival probability is equal to the product of the first six conditional probabilities.

```{r t6}
km$n[6] <-  8
km$d[6] <-  2 
km$c[6] <-  0
km$p[6] <- 1-km$d[6]/km$n[6]
km$s[6] <- km$p[1]*km$p[2]*km$p[3]*km$p[4]*km$p[5]*km$p[6]
km
```

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, 2+, 2+, 2, 2, 
3+, 3, 4+, 4, 5+, 5+, 5+, 5+, 5+, 5, 5, 5, 6, 6, **7,** 
10, 10, 12, 12, 13

At t7=7, there are n7=6 at risk, and you have d7=1 failure and c7=0 censored observations. The survival probability is equal to the product of the first seven conditional probabilities.

```{r t7}
km$n[7] <-  6
km$d[7] <-  1 
km$c[7] <-  0
km$p[7] <- 1-km$d[7]/km$n[7]
km$s[7] <- km$p[1]*km$p[2]*km$p[3]*km$p[4]*km$p[5]*km$p[6]*km$p[7]
km
```

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, 2+, 2+, 2, 2, 
3+, 3, 4+, 4, 5+, 5+, 5+, 5+, 5+, 5, 5, 5, 6, 6, 7,
**10, 10,** 12, 12, 13

At t8=10, there are n8=5 at risk, and you have d8=2 failures and c8=0 censored observations. The survival probability is equal to the product of the first eight conditional probabilities.

```{r t8}
km$n[8] <-  5
km$d[8] <-  2 
km$c[8] <-  0
km$p[8] <- 1-km$d[8]/km$n[8]
km$s[8] <- km$p[1]*km$p[2]*km$p[3]*km$p[4]*km$p[5]*km$p[6]*km$p[7]*km$p[8]
km
```

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, 2+, 2+, 2, 2, 
3+, 3, 4+, 4, 5+, 5+, 5+, 5+, 5+, 5, 5, 5, 6, 6, 7,
10, 10, **12, 12,** 13

At t9=12, there are n9=3 at risk, and you have d9=2 failures and c9=0 censored observations. The survival probability is equal to the product of the first nine conditional probabilities.

```{r t9}
km$n[9] <-  3
km$d[9] <-  2 
km$c[9] <-  0
km$p[9] <- 1-km$d[9]/km$n[9]
km$s[9] <- km$p[1]*km$p[2]*km$p[3]*km$p[4]*km$p[5]*km$p[6]*km$p[7]*km$p[8]*km$p[9]
km
```

1+, 1+, 1+, 1+, 1+, 1+, 1+, 1+, 1, 1, 2+, 2+, 2, 2, 
3+, 3, 4+, 4, 5+, 5+, 5+, 5+, 5+, 5, 5, 5, 6, 6, 7,
10, 10, 12, 12, **13**

At t10=12, there is n10=1 at risk, and you have d10=1 failure and c10=0 censored observations. The survival probability is equal to the product of all ten conditional probabilities.

```{r t10}
km$n[10] <-  1
km$d[10] <-  1 
km$c[10] <-  0
km$p[10] <- 1-km$d[10]/km$n[10]
km$s[10] <- km$p[1]*km$p[2]*km$p[3]*km$p[4]*km$p[5]*km$p[6]*km$p[7]*km$p[8]*km$p[9]* km$p[10]
km
```

Save everything for possible re-use.

```{r save-everything}
save.image("../../data/homework1.RData")
```

Well, that was a lot of work, but it was worth it.

Let's input the data to check our work.

```{r check}
t <- c(1:7, 10, 12, 13)

t <- c(
  1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 
  3, 3, 4, 4, 5, 5, 5, 5, 5, 5, 5, 5, 6, 6,
  7, 10, 10, 12, 12, 13)

i <- c(
  0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 0, 0, 1, 1, 
  0, 1, 0, 1, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1,
  1, 1, 1, 1, 1, 1)

catheter <- data.frame(t=t, i=i)
catheter_surv <- Surv(catheter$t, catheter$i)
catheter_km <- summary(survfit(catheter_surv~1))
catheter_km
data.frame(our.calc=km$s, r.calc=catheter_km$surv)
```

They match. Hooray!
