---
title: "survival lecture 5"
author: "Steve Simon"
date: "May 27, 2018"
output: html_document
---

Lecture 5. Parametric models. Parametric models provide an alternative analysis to the Cox proportional hazards model. You'll compare the hazard function for various popular survival distributions and understand the advantages and disadvantages of a parametric approach to survival. You'll fit parametric models and interpret the coefficients.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```


```{r read-whas100}
fn <- "~/survival-models/bin/whas100.RData" 
load(fn)
whas100$time_yrs <- whas100$lenfol / 365.25
head(whas100)
whas100_surv <- Surv(whas100$time_yrs, whas100$fstat=="Dead")
```

Before you fit the exponential and Weibull regression models, fit a Kaplan-Meier curve.

```{r km, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
km_overall <- survfit(Surv(whas100$time_yrs, whas100$fstat)~1)
exp_null_model <- survreg(whas100_surv~1, dist="exponential")
summary(exp_null_model)
pct <- (1:999)/1000

exp_null_model %>%
  predict(
    newdata=data.frame(one=1),
    type="quantile",
    p=pct)                     %>%
  data.frame                   %>%
  set_names("time")            %>%
  mutate(estimate=rev(pct))    %>%
  mutate(grp="Exponential")    %>%
  filter(time<=10)             -> exp_survival

km_overall                     %>%
  tidy                         %>%
  mutate(estimate=1-estimate)  %>%
  mutate(grp="Kaplan-Meier")   %>%
  bind_rows(exp_survival)      -> combined_survivals

combined_survivals             %>%
  ggplot(aes(time, estimate))   +
    expand_limits(y=0)          +
    geom_step(aes(color=grp))

```

You can also compare the cumulative hazard functions.

```{r cumhaz, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
combined_survivals                %>%
  mutate(c_hazard=-log(estimate)) %>%
  ggplot(aes(time, c_hazard))      +
    geom_step(aes(color=grp))
```

```{r exponential-regression, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
whas100$age_c=whas100$age-mean(whas100$age)
whas100$bmi_c=whas100$bmi-mean(whas100$bmi)
whas100$i_female=as.numeric(whas100$gender=="Female")
exp_model <- survreg(
  whas100_surv~i_female*age_c+bmi_c, 
  data=whas100, 
  dist="exponential")
summary(exp_model)
cox_model <- coxph(
  whas100_surv~i_female*age_c+bmi_c, 
  data=whas100)
summary(cox_model)
exp_coef <- tidy(exp_model)
bmi_coef <- exp(exp_coef[exp_coef$term=="bmi_c", "estimate"])
age_m_coef <- exp(exp_coef[exp_coef$term=="age_c", "estimate"])
age_f_coef <- exp(exp_coef[exp_coef$term=="age_c", "estimate"]+
  exp_coef[exp_coef$term=="i_female:age_c", "estimate"])
```

Each unit increase in BMI shifts the survival percentiles by a factor of exp(`r round(log(bmi_coef), 3)`) = `r round(bmi_coef, 3)`. A five unit shift in BMI shifts the survival percentiles by `r round(bmi_coef, 3)`^5 = `r round(bmi_coef^5, 3)`. 

Because of the interaction, the effect of age is different for men and women. For men, a one year increase in age will shift the survival percentiles by a factor of
exp(`r round(exp_coef[exp_coef$term=="age_c", "estimate"], 3)`) = `r round(age_m_coef, 3)`. 

For women, a one year increase in age will shift the survival percentiles by a factor of exp(`r round(exp_coef[exp_coef$term=="age_c", "estimate"], 3)` + `r round(exp_coef[exp_coef$term=="i_female:age_c", "estimate"], 3)`) = `r round(age_f_coef, 3)`.

This may look quite different than the model we used for Kaplan-Meier curves and the Cox proportional hazards model, but it actually is not. The hazard function is 

$h(t, x, \beta_0, \beta_1)=e^{-(\beta_0+\beta_1 x)}$.

Notice that the hazard is constant with respect to t. The baseline hazard, the hazard when X=0 is

$h_0(t)=e^{-\beta_0}$

The hazard ratio for a subject with $x=x_1$ compared to a subject with $x=x_2$ is

$e^{-\beta_1 (x_1-x_2)}$


```{r null-weibull, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
weibull_null_model <- survreg(whas100_surv~1, dist="weibull")
summary(weibull_null_model)

pct <- (1:999)/1000

weibull_null_model %>%
  predict(
    newdata=data.frame(one=1),
    type="quantile",
    p=pct)                     %>%
  data.frame                   %>%
  set_names("time")            %>%
  mutate(estimate=rev(pct))    %>%
  mutate(grp="Weibull")        %>%
  filter(time<=10)             -> weibull_survival

combined_survivals             %>%
  bind_rows(weibull_survival)  -> combined_survivals

combined_survivals             %>%
  ggplot(aes(time, estimate))   +
    expand_limits(y=0)          +
    geom_step(aes(color=grp))

```

```{r weibull-regression}
weibull_model <- survreg(
  whas100_surv~i_female*age_c+bmi_c, 
  data=whas100, 
  dist="weibull")
summary(weibull_model)
coef_w <- tidy(weibull_model)
coef_w
log_scale <- coef_w[coef_w$term=="Log(scale)", "estimate"]
std_error <- coef_w[coef_w$term=="Log(scale)", "std.error"]
ci_lower <- exp(log_scale-1.96*std_error)
ci_upper <- exp(log_scale+1.96*std_error)
```

The 95% confidence interval for the scale parameter is 
exp(`r round(log_scale, 3)` +/- 1.96*`r round(std_error, 3)`) or
`r round(ci_lower, 3)` to 
`r round(ci_upper, 3)`.

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/survival-lecture-5.RData")
```