---
title: "survival lecture 4"
author: "Steve Simon"
date: "April 8, 2018"
output: html_document
---

Lecture 4. Model fitting and diagnostics for the Cox model. In this lecture, you will work with more complex forms of the Cox model with multiple predictor variables. You'll include covariates in the Cox model to produce risk adjusted survival curves. You will also assess the underlying assumptions of the Cox model, particularly the assumption of proportional hazards.

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

## Cox multivariate regression

Recall that we ran a log rank test for gender using the WHAS100 data file. Let's calculate a Cox regression model for the same data and compare it to the log rank test.

```{r read}
fn <- "../../data/whas100.RData" 
load(fn)
head(whas100)
fn <- "../../data/whas500.RData" 
load(fn)
whas500$time_yrs <- whas500$lenfol/365.25
head(whas500)
```

You should examine your independent variables one at a time before putting them all in a multivariate model.

```{r one-at-a-time, fig.width=4.5, fig.height=2.5}
cox_age <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age,
    data=whas500)
summary(cox_age)
cox_gender <- 
  coxph(Surv(time_yrs, fstat=="Dead")~gender,
    data=whas500)
summary(cox_gender)
n <- data.frame(gender=factor(1:2, labels=c("Male", "Female")))
cox_predict <- survfit(cox_gender, newdata=n)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  cox_predict$time,
  cox_predict$surv[, 2], 
  type="s", ylim=c(0,1))
lines(
  cox_predict$time, 
  cox_predict$surv[, 1], 
  type="s", lty="dotted")
five_year <- which.min(abs(cox_predict$time-5))
y <- cox_predict$surv[five_year, ]+0.05
text(5, y, c("M", "F"))
cox_mv <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age+gender,
    data=whas500)
summary(cox_mv)
```

The effect of gender disappears when you include age in the model.

```{r boxplots, fig.width=4.5, fig.height=2.5}
overall_mean <- mean(whas500$age)
gender_means <- tapply(whas500$age, whas500$gender, mean)
age_diff <- gender_means[2]-gender_means[1]
overall_mean
gender_means
ggplot(whas500, aes(x=gender, y=age)) +
  geom_boxplot()
coef_mv <- tidy(cox_mv)
coef_mv
hr_age <- exp(coef_mv[coef_mv$term=="age", "estimate"])
extrapolated_hr <- hr_age^age_diff
```

Notice that there is a `r round(age_diff, 1)` year difference between the average ages of men and women. The hazard ratio for age, `r round(hr_age, 3)`, can get extrapolated to a `r round(age_diff, 1)` year difference by exponentiating. That is `r round(hr_age, 3)`^`r round(age_diff, 1)` = `r round(extrapolated_hr, 2)` which is actually larger than the hazard ratio that we saw for the unadjusted model with just gender.

```{r unadjusted-and-adjusted, fig.width=4.5, fig.height=2.5}
n_unadjusted <- 
  data.frame(
    age=gender_means, 
    gender=factor(1:2, labels=c("Male", "Female")))
predict_unadjusted <- survfit(cox_mv, newdata=n_unadjusted)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_unadjusted$time,
  predict_unadjusted$surv[, 2], 
  type="s", ylim=c(0,1))
lines(
  predict_unadjusted$time, 
  predict_unadjusted$surv[, 1], 
  type="s", lty="dotted")
five_year <- which.min(abs(predict_unadjusted$time-5))
y <- predict_unadjusted$surv[five_year, ]+0.05
text(5, y, c("M", "F"))

n_adjusted <- 
  data.frame(
    age=overall_mean,
    gender=factor(1:2, labels=c("Male", "Female")))
predict_adjusted <- survfit(cox_mv, newdata=n_adjusted)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_adjusted$time,
  predict_adjusted$surv[, 2], 
  type="s", ylim=c(0,1))
lines(
  predict_adjusted$time, 
  predict_adjusted$surv[, 1], 
  type="s", lty="dotted")
five_year <- which.min(abs(predict_adjusted$time-5))
y <- predict_adjusted$surv[five_year, ]+0.05
text(5, y, c("M", "F"))
```

```{r compare-age-and-bmi, fig.width=4.5, fig.height=2.5}
whas500$i_female <- as.numeric(whas500$gender=="Female")
cox_bmi <- 
  coxph(Surv(time_yrs, fstat=="Dead")~bmi,
    data=whas500)
summary(cox_bmi)
n_bmi <- data.frame(bmi=c(20, 30, 40))
predict_bmi <- survfit(cox_bmi, newdata=n_bmi)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_bmi$time,
  predict_bmi$surv[, 1], 
  type="s", ylim=c(0,1))
lines(
  predict_bmi$time, 
  predict_bmi$surv[, 2], 
  type="s", lty="dotted")
lines(
  predict_bmi$time, 
  predict_bmi$surv[, 3], 
  type="s", lty="dashed")
five_year <- which.min(abs(predict_bmi$time-5))
y <- predict_bmi$surv[five_year, ]+0.05
text(5, y, c(20, 30, 40))
ggplot(data=whas500) +
  geom_boxplot(aes(x=gender, y=bmi))
ggplot(data=whas500) +
  geom_point(aes(x=age, y=bmi))
cox_bmi_adj <- 
  coxph(Surv(time_yrs, fstat=="Dead")~bmi+age+i_female,
    data=whas500)
summary(cox_bmi_adj)
n_bmi <- data.frame(age=overall_mean, i_female=1, bmi=c(20, 30, 40))
predict_bmi <- survfit(cox_bmi_adj, newdata=n_bmi)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_bmi$time,
  predict_bmi$surv[, 1], 
  type="s", ylim=c(0,1))
lines(
  predict_bmi$time, 
  predict_bmi$surv[, 2], 
  type="s", lty="dotted")
lines(
  predict_bmi$time, 
  predict_bmi$surv[, 3], 
  type="s", lty="dashed")
five_year <- which.min(abs(predict_bmi$time-5))
y <- predict_bmi$surv[five_year, ]+0.05
text(5, y, c(20, 30, 40))
```

Look at the interaction.

```{r int}
cox_int1 <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age*i_female,
    data=whas500)
summary(cox_int1)
whas500$age_centered <- whas500$age-overall_mean
cox_int2 <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age_centered*i_female,
    data=whas500)
summary(cox_int2)
coef_int2 <- tidy(cox_int2)$estimate
hr <- exp(coef_int2[2]+coef_int2[3]*((40:90)-overall_mean))
plot(40:90, hr, type="l")
abline(h=1)

```

You can use a sequence of Wald tests to compare different models for survival.

```{r wald}
cox_v1 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    gender,
      data=whas500)

cox_v2 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    age+gender,
      data=whas500)                

cox_v3 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    bmi+age+gender,
      data=whas500)

cox_v1                             %>%
  tidy                             %>%
  mutate(lab="gender only")        -> wald_v1

cox_v2                             %>%
  tidy                             %>%
  mutate(lab="age, gender")        -> wald_v2

cox_v3                             %>%
  tidy                             %>%
  mutate(lab="bmi, age, gender")   -> wald_v3

wald_v1                            %>%
  bind_rows(wald_v2)               %>%
  bind_rows(wald_v3)               %>%  
  select(lab, term, estimate, p.value) -> wald_all
wald_all
```

You use the log partial likelihood and/or the AIC (Akaike Information Criteria) to compare models of different complexity.

AIC = -2 log Likelihood + 2 k.

AIC = -2 log Likelihood + log(n) k.

```{r aic}
cox_v1                           %>%
  glance                         %>%
  mutate(lab="gender only")      -> aic1

cox_v2                           %>%
  glance                         %>%
  mutate(lab="gender, age")      -> aic2

cox_v3                           %>%
  glance                         %>%
  mutate(lab="gender, age, bmi") -> aic3

aic1                             %>%
  bind_rows(aic2)                %>%
  bind_rows(aic3)                %>%
  select(lab, logLik, AIC, BIC)  -> aic_all
aic_all
```

There are several residuals available for Cox regression. The Martingale residual is defined as

$M(t_i) = \delta_i - H_0(t_i) e^{X\beta}$

where $\delta_i$ = 0 if censored, 1 if dead.

```{r residuals, fig.width=4.5, fig.height=2.5}
whas500$r <- residuals(cox_v3)
par(mar=c(2.6, 2.6, 0.6, 0.6))
ggplot(whas500, aes(age, r)) +
  geom_point()
ggplot(data=whas500, aes(age, r)) +
  geom_point() +
  geom_smooth(se=TRUE)
ggplot(data=whas500, aes(hr, r)) +
  geom_point() +
  geom_smooth(se=TRUE)
ggplot(data=whas500, aes(sysbp, r)) +
  geom_point() +
  geom_smooth(se=TRUE)
ggplot(data=whas500, aes(diasbp, r)) +
  geom_point() +
  geom_smooth(se=TRUE)
ggplot(data=whas500, aes(cvd, r)) +
  geom_boxplot()
ggplot(data=whas500, aes(afb, r)) +
  geom_boxplot()
ggplot(data=whas500, aes(sho, r)) +
  geom_boxplot()
ggplot(data=whas500, aes(chf, r)) +
  geom_boxplot()
ggplot(data=whas500, aes(av3, r)) +
  geom_boxplot()
ggplot(data=whas500, aes(miord, r)) +
  geom_boxplot()
ggplot(data=whas500, aes(mitype, r)) +
  geom_boxplot()
```

```{r spline, fig.width=4.5, fig.height=2.5}
cox_pspline3 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    bmi+pspline(age, df=3)+gender,
      data=whas500)

cox_pspline8 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    bmi+pspline(age, df=8)+gender,
      data=whas500)

terms_pspline3 <- predict(cox_pspline3, type="terms")
terms_pspline8 <- predict(cox_pspline8, type="terms")
par(mar=c(2.6, 2.6, 0.6, 0.6))
o <- order(whas500$age)
plot(whas500$age[o], terms_pspline3[o ,2], type="l")
plot(whas500$age[o], terms_pspline8[o ,2], type="l")
glance(cox_bmi_adj) %>%
  bind_rows(glance(cox_pspline3)) %>%
  bind_rows(glance(cox_pspline8)) %>%
  mutate(lab=c(
    "linear (df=1)",
    "spline (df=3)",
    "spline (df=8)"))             %>%
  select(lab, logLik, AIC, BIC)   -> compare_splines
compare_splines
```

Testing.

```{r other-models}
coxph(Surv(time_yrs, fstat=="Dead")~hr, data=whas500)
coxph(Surv(time_yrs, fstat=="Dead")~sysbp, data=whas500)
coxph(Surv(time_yrs, fstat=="Dead")~diasbp, data=whas500)
coxph(Surv(time_yrs, fstat=="Dead")~cvd, data=whas500)
coxph(Surv(time_yrs, fstat=="Dead")~afb, data=whas500)
coxph(Surv(time_yrs, fstat=="Dead")~sho, data=whas500)
coxph(Surv(time_yrs, fstat=="Dead")~chf, data=whas500)
coxph(Surv(time_yrs, fstat=="Dead")~av3, data=whas500)
coxph(Surv(time_yrs, fstat=="Dead")~miord, data=whas500)
coxph(Surv(time_yrs, fstat=="Dead")~mitype, data=whas500)
```

Save everything for possible later re-use.

```{r save-everything}
save.image("../../data/class4.RData")
```