---
title: "supplement 7"
author: "Steve Simon"
date: "June 2, 2018"
output: html_document
---

This file creates various formulas and graphs needed to illustrate the underlying theory behind the data analyses in survival-lecture-7.Rmd. I am sharing this for those who are curious, but you are not responsible for learning or using the code shown in this supplement.

```{r render, eval=FALSE, echo=FALSE}
# run the commands in this section to store the output
# in the results folder.
library(rmarkdown)
f <- "~/survival-models/src/supplement-7.Rmd"
o <- "~/survival-models/results"
render(f, output_dir=o)
```

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May.

```{r load-libraries}
library(broom)
library(coxme)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

## Interval censored data

An event is said to be interval censored in the interval (a,b) if you know that the event did not occur before time a and that it did occur before time b.

A right censored observation is equivalent to interval censoring on the interval (a, $\infty$).

A left censored observation is equivalent to interval censoring on the interval (0, b).

The likelihood in most settings is the product of the density functions, but with interval censoring, you replace the density by

$\int_a^b f(t; \beta) = S(a; \beta)-S(b; \beta)$

## Left truncation

In most studies, patients are considered part of the risk set from time 0 to the time that they either die or are censored. But sometimes patients start contributing information about survival only from a certain point forward. This is left truncation.

Here's an example from

math.usu.edu/jrstevens/biostat/projects2013/pres_LeftTruncation.pdf

```{r psychiatric, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6), las=1)
fn <- "~/survival-models/bin/psychiatric.RData"
load(fn)
head(psychiatric)
psychiatric %<>% arrange(age, age2)
n <- dim(psychiatric)[1]
n_atrisk <- rep(NA, 80)
for (i in 1:80) {
  n_atrisk[i] <- sum((psychiatric$age<=i) - (psychiatric$age2<=i))
}
plot(c(0, 80), c(1,n), type="n")
segments(psychiatric$age, 1:n, psychiatric$age2, 1:n)
plot(1:80, n_atrisk, type="s", ylim=c(0,25))
```

## Sandwich estimate

You can account for cluster effects by modifying the variance-covariance of the coefficients. For the normal model without frailty effects, the information matrix is

$I(\beta) = \frac{\partial^2 L_p}{(\partial \beta)^2}$

and the estimated variance covariance matrix, $\hat{V}$, is

$\hat{V} = I(\hat{\beta})^{-1}$

The robust sandwich estimator (similar to the sandwich estimator in Generalized Estimating Equations) is

$\hat{R} = \hat{V} (\hat{L}'\hat{L}) \hat{V}$

where $\hat{L}$ is the vector of Schoenfeld residuals. The middle of this sandwich, the $\hat{L}'\hat{L}$, adjusts the variance covariance matrix to account for the correlation within clusters.

## Frailty model

The frailty model introduces a random effect, called a frailty term, $z_f$. Recall that the proportional hazards model assumes that the hazard function of a given patient,

$h(t, X, \beta) = h_0(t) e^{X \beta}$

The frailty model multiplies all the hazards within a family f by $z_f$. In other words,

$h(t, X, \beta) = z_f h_0(t) e^{X \beta}$

Typically, $z_f$ is given a gamma distribution with a mean of 1 and a variance of $\theta$.

```{r graph-gamma, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
alpha <- 10
x <- seq(0.001, 4, length=1000)
plot(x, dgamma(x, alpha, rate=alpha), 
  type="l", xlim=c(0, 4), ylim=c(0, 1.3))

alpha <- 5
plot(x, dgamma(x, alpha, rate=alpha), 
  type="l", xlim=c(0, 4), ylim=c(0, 1.3))

alpha <- 2
plot(x, dgamma(x, alpha, rate=alpha), 
  type="l", xlim=c(0, 4), ylim=c(0, 1.3))
```

```{r three-models, fig.width=4.5, fig.height=2.5}
# kidney is a data set found in the survival library.
par(mar=c(2.6, 2.6, 0.6, 0.6))
plot_sequence <- function(i) {
  c1 <- "darkred"
  c2 <- "darkgreen"
  kidney2$lb <- ifelse(kidney2$status==1, kidney2$s, "C")
  kidney2                %>%
    filter(id==i)        %>%
    filter(s==1)         %>%
    select(time, lb)     -> t1
  kidney2                %>%
    filter(id==i)        %>%
    filter(s==2)         %>%
    select(time, lb)     -> t2
  t2$time <- ifelse(t1$lb=="C", NA, t2$time)
  tmax <- sum(t1$time,t2$time, na.rm=TRUE)
  plot(c(0, 1.4*tmax), c(10, 50), type="n", axes=FALSE)
  axis(side=1, at=c(0, t1$time, t2$time, tmax))
  segments(0, 50, t1$time, 50, col=c1)
  segments(t1$time, 50, tmax, 50, col=c2)
  text(t1$time, 50, t1$lb, col=c1)
  text(tmax,    50, t2$lb, col=c2)
  text(1.2*tmax, 45, "start", adj=1)
  text(1.4*tmax, 45, "stop", adj=1)

  segments(0, 40, t1$time, 40, col=c1)
  segments(0, 36, t2$time, 36, col=c2)
  text(t1$time, 40, t1$lb, col=c1)
  text(t2$time, 36, t2$lb, col=c2)
  text(1.2*tmax, 40, 0, col=c1, adj=1)
  text(1.4*tmax, 40, t1$time, col=c1, adj=1)
  text(1.2*tmax, 36, 0, col=c2, adj=1)
  text(1.4*tmax, 36, t2$time, col=c2, adj=1)
  
  segments(0, 30, t1$time, 30, col=c1)
  segments(t1$time, 26, tmax, 26, col=c2)
  text(t1$time, 30, t1$lb, col=c1)
  text(tmax, 26, t2$lb, col=c2)
  text(1.2*tmax, 30, 0, col=c1, adj=1)
  text(1.4*tmax, 30, t1$time, col=c1, adj=1)
  text(1.2*tmax, 26, t1$time, col=c2, adj=1)
  text(1.4*tmax, 26, tmax, col=c2, adj=1)

  segments(0, 20, t1$time, 20, col=c1)
  segments(0, 16, tmax, 16, col=c2)
  text(t1$time, 20, t1$lb, col=c1)
  text(tmax, 16, t2$lb, col=c2)
  text(1.2*tmax, 20, 0, col=c1, adj=1)
  text(1.4*tmax, 20, t1$time, col=c1, adj=1)
  text(1.2*tmax, 16, 0, col=c2, adj=1)
  text(1.4*tmax, 16, tmax, col=c2, adj=1)
}
plot_sequence(30)
```

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/supplement-7.RData")
```