---
title: "homework answers 4"
author: "Steve Simon"
date: "May 17, 2018"
output: html_document
---

1. Open the WHAS500 data set in the software program of 
your choice.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
fn <- "../data/whas500.RData"
load(fn)
head(whas500)
```

a. Calculate a Cox regression model systolic blood 
pressure (sysbp) by itself ...

```{r cox-sysbp}
whas500$time_yrs <- whas500$lenfol / 365.25
cox_sysbp <- coxph(
  Surv(time_yrs, fstat=="Dead")~sysbp,
    data=whas500)
summary(cox_sysbp)
```

... and then adjusted for gender and age.

```{r cox-sysbp-adj}
i_female <- as.numeric(whas500$gender=="Female")
cox_sysbp_adj <- coxph(
  Surv(time_yrs, fstat=="Dead")~sysbp+age+i_female,
    data=whas500)
summary(cox_sysbp_adj)
```

Is there evidence that systolic blood pressure
shows any trends with respect to gender or age?

```{r trends}
ggplot(data=whas500, aes(age, sysbp)) +
  geom_point()
ggplot(data=whas500, aes(gender, sysbp)) +
  geom_boxplot()
```

Calculate the unadjusted survival curves for patients
with systolic blood pressures of 120, 140, and 160.

```{r unadjusted-survival}
n_unadjusted <- 
  data.frame(
    sysbp=c(120, 140, 160))
predict_unadjusted <- survfit(cox_sysbp, newdata=n_unadjusted)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_unadjusted$time,
  predict_unadjusted$surv[, 3], 
  type="s", ylim=c(0,1))
lines(
  predict_unadjusted$time, 
  predict_unadjusted$surv[, 2], 
  type="s", lty="dotted")
lines(
  predict_unadjusted$time, 
  predict_unadjusted$surv[, 1], 
  type="s", lty="dashed")
```

Then recalculate these survival curves with age set to the 
overall average age, and to a population that is 30%
female. Interpret your results.

```{r adjusted-survival}
n_adjusted <- 
  data.frame(
    i_female=0.3,
    age=mean(whas500$age),
    sysbp=c(120, 140, 160))
predict_adjusted <- survfit(cox_sysbp_adj, newdata=n_adjusted)
par(mar=c(2.1, 2.1, 0.6, 0.6))
plot(
  predict_adjusted$time,
  predict_adjusted$surv[, 3], 
  type="s", ylim=c(0,1))
lines(
  predict_adjusted$time, 
  predict_adjusted$surv[, 2], 
  type="s", lty="dotted")
lines(
  predict_adjusted$time, 
  predict_adjusted$surv[, 1], 
  type="s", lty="dashed")
```

b. Calculate cubic spline model for systolic blood
pressure with four degrees of freedom.

```{r calculate-spline}
cox_pspline4 <- coxph(
  Surv(time_yrs, fstat=="Dead") ~ 
    pspline(sysbp, df=4),
      data=whas500)
cox_pspline4
```

Plot this spline and offer an informal assessment as
to whether your spline function deviates markedly 
from a linear relationship.

```{r plot-spline-risk}
cox_pspline4 %>%
  augment(data=whas500, type.predict="risk") %>%
  ggplot(aes(x=sysbp, y=.fitted)) +
    geom_point()
```

c. Calculate the Martingale residuals from your Cox
model with a linear term for systolic blood pressure
and for age and a term for gender. Plot these residuals
versus diastolic blood pressure. Offer an informal
assessment as to whether diastolic blood pressure adds
any information to a model already including systolic
blood pressure. Repeat this residual plot analysis using
myocardial infection type (mitype).

```{r martingale}
i_female <- as.numeric(whas500$gender=="Female")
cox_sysbp_adj <- coxph(
  Surv(time_yrs, fstat=="Dead")~sysbp+age+i_female,
    data=whas500)
# I am re-including the code for the model here for clarity.
martingale_residuals <-
  residuals(cox_sysbp_adj, type="martingale")
ggplot(whas500, aes(x=diasbp, y=martingale_residuals)) +
  geom_point() +
  geom_smooth()
ggplot(whas500, aes(x=mitype, y=martingale_residuals)) +
  geom_boxplot()
```