---
title: "survival lecture 6"
author: "Steve Simon"
date: "May 20, 2018"
output: html_document
---

Lecture 6. Time varying covariates in a Cox model. Time varying covariates allow you to account for non-proportional hazards and can model settings where patients switch from one therapy to another. You will code data for time-varying covariates, fit time-varying models, and interpret the results.

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

## The Kaplan-Meier plot and proportional hazards.

```{r read}
fn <- "~/survival-models/bin/whas500.RData"
load(fn)
fn <- "~/survival-models/bin/heroin.RData"
load(fn)
head(heroin)
table(heroin$status)
summary(heroin)
```

```{r km-plots, fig.width=4.5, fig.height=2.5}
heroin                                      %$%
  Surv(time_yrs, status=="departed")        -> heroin_surv

heroin                                      %$%
  survfit(heroin_surv~clinic)               %>%
  tidy                                      -> km_clinic

km_clinic                                   %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()

heroin                                      %$%
  survfit(heroin_surv~prison_record)        %>%
  tidy                                      -> km_prison

km_prison                                   %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()

heroin                                      %$%
  survfit(heroin_surv~dose_group)           %>%
  tidy                                      -> km_dose

km_dose                                     %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()
```

## Complementary log-log plot

```{r log-log, fig.width=4.5, fig.height=2.5}
km_clinic                             %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    expand_limits(y=0)                 +
    geom_step()

km_prison                             %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    expand_limits(y=0)                 +
    geom_step()

km_dose                               %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    expand_limits(y=0)                 +
    geom_step()
```

## Schoenfeld residuals.

```{r schoenfeld, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 4.1, 0.6, 0.6))
cox_mv <- coxph(
  heroin_surv~clinic+i_prison+methadone_dose,
  data=heroin)
cox_schoenfeld <- cox.zph(cox_mv)
print(cox_schoenfeld)
plot(cox_schoenfeld)
```

## Stratified analysis

If one of the variables in your model is a nuisance variable, but it could possibly be the cause of the violation of the proportional hazards assumption, you can fit fix this by fitting a separate baseline hazard for each level of your nuisance variable.

```{r stratified-analysis, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
cox_strata <- coxph(
  heroin_surv ~
    i_prison +
    methadone_dose +
  strata(clinic),
  data=heroin)
print(cox_strata)
summary(cox_strata)

n <- data.frame(
  methadone_dose=mean(heroin$methadone_dose),
  clinic=c(1, 1, 2, 2),
  i_prison=c(0, 1, 0, 1))

cox_predict <- survfit(cox_strata, newdata=n)
plot(
  cox_predict$time[1:147],
  cox_predict$surv[1:147],
  type="s", ylim=c(0,1))
lines(
  cox_predict$time[148:294], 
  cox_predict$surv[148:294], 
  type="s", lty="dotted")
plot(
  cox_predict$time[295:362], 
  cox_predict$surv[295:362], 
  type="s", ylim=c(0,1))
lines(
  cox_predict$time[363:430], 
  cox_predict$surv[363:430], 
  type="s", lty="dotted")
```

## Time-varying covariates

```{r time-transfer, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
tt_clinic <- coxph(
  heroin_surv~
    clinic +
    tt(clinic) +
    i_prison +
    methadone_dose,
  tt=function(x, t, ...) x*t,
  data=heroin)
tt_clinic
summary(tt_clinic)
tt_coef <- tidy(tt_clinic)
beta_clinic <- tt_coef[tt_coef$term=="clinic", "estimate"]
beta_interaction <- tt_coef[tt_coef$term=="tt(clinic)", "estimate"]
t <- seq(0, 2.5, by=0.5)
log_hr <- beta_clinic+t*beta_interaction
hr <- exp(log_hr)
round(
  data.frame(
    beta_clinic, 
    beta_interaction, 
    t, 
    log_hr,
    hr),
  3)
plot(t, hr, ylim=c(0, 1.1))
```

## Stanford Heart Transplant study

The Stanford Heart Transplant study has an interesting application of time-varying covariates.

```{r read-heart, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6), las=1)
ids <- 21:30
vlist <- c(
  "accept.dt",
  "tx.date",
  "fu.date",
  "fustat",
  "transplant"
)
jasa [ids, vlist]
t0 <- as.Date("1969-01-01")
t1 <- jasa$accept.dt[ids] - t0
t2 <- jasa$tx.date[ids] - t0
t3 <- jasa$fu.date[ids] - t0
censor <- ifelse(jasa$fustat[ids]==1, "D", "C")
plot(c(0, max(t3)), c(1,10), type="n", axes=FALSE)
axis(side=1, at=(0:5)*365.25, labels=1969:1974)
axis(side=2, at=1:10, label=ids)
box()
segments(t1, 1:10, t3, 1:10)
text(t2, (1:10)+0.2, "T", col="darkgreen", cex=0.5)
text(t3, 1:10, censor, col="red", adj=0, cex=0.5)
data.frame(
  id=ids,
  time1=t2-t1, 
  time2=t3-t1, 
  event=censor,
  transplant=jasa$transplant[ids])
jasax <- jasa1
row.names(jasax) <- NULL
col.names <- c("id", "start", "stop", "event", "transplant")
jasax[jasax$id %in% ids, col.names]
```

```{r naive-model}
cox_naive <- coxph(
  Surv(futime, fustat) ~ transplant + age + surgery,
    data=jasa)

print(cox_naive)

cox_time_varying <- coxph(
  Surv(start, stop, event, type="counting") ~
    transplant+age+surgery,
  data=jasa1)

print(cox_time_varying)
```

Page 221 of Hosmer, Lemeshow, and May describes the grace data set, which you can use for time-varying covariates.


```{r read-grace, eval=FALSE}
fn <- "~/survival-models/data/wiley/GRACE1000.dat" 
grace <- read.table(fn, header = FALSE) 
names(grace) <- c(
  "id",
  "days",
  "death",
  "revasc",
  "revascdays",
  "los",
  "age",
  "sysbp",
  "stchange")
head(grace)
```

```{r extras, eval=FALSE}
whas500 %$% Surv(time_yrs, fstat=="Dead") -> whas500_surv
bmi_cuts <- c(0, 25, 30, 35, 99)
whas500                                     %>%
  mutate(bmi_categories=cut(bmi, bmi_cuts)) %$%
  survfit(whas500_surv~bmi_categories)      %>%
  tidy                                      -> km_bmi
km_bmi                                      %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()

age_cuts <- c(0, 60, 70, 80, 999)
whas500                                     %>%
  mutate(age_categories=cut(age, age_cuts)) %$%
  survfit(whas500_surv~age_categories)      %>%
  tidy                                      -> km_age
km_age                                      %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()

hr_cuts <- c(0, 70, 85, 100, 999)
whas500                                    %>%
  mutate(hr_categories=cut(hr, hr_cuts))   %$%
  survfit(whas500_surv~hr_categories)      %>%
  tidy                                      -> km_hr
km_hr                                       %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()

diasbp_cuts <- c(0, 60, 70, 80, 999)
whas500                                     %>%
  mutate(
    diasbp_categories=
      cut(diasbp, diasbp_cuts))             %$%
  survfit(whas500_surv~diasbp_categories)   %>%
  tidy                                      -> km_diasbp
km_diasbp                                   %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()

whas500                                     %$%
  survfit(whas500_surv~gender)              %>%
  tidy                                      -> km_gender
km_gender                                   %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()

whas500                                     %$%
  survfit(whas500_surv~chf)                 %>%
  tidy                                      -> km_chf
km_chf                                      %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()
km_bmi                                %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()

km_age                                %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()

km_hr                                 %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()

km_diasbp                             %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()

km_gender                             %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()

km_chf                                %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()
cox_mv <- coxph(
  Surv(time_yrs, fstat=="Dead") ~
    bmi +
    age +
    gender,
  data=whas500)
cox_schoenfeld <- cox.zph(cox_mv)
print(cox_schoenfeld)
plot(cox_schoenfeld)

```

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/survival-lecture-6.RData")
```