---
title: "Read and code data files"
author: "Steve Simon"
date: "April 24, 2018"
output: html_document
---

This program reads in several text files, makes minor updates, and adds documentation to them. It uses an R library, kutils, that stores the update information in an easy to read spreadsheet or comma separated text file.

## The kutils template

The first step is to create a template with the existing information about the data frame. The keyTemplate function in kutils does this for you.

### Template variable names

You can review the template to find the current name of every variable in the data frame under "name_old". The template has a column right next to it, "name_new" that you can use to rename the variable to something more close to your liking. 

### Template variable types

It stores the current class of the variable (integer, numeric, character, etc.) under "class_old" and has a second column right next to it, "class_new". You can change the class (e.g., from numeric to character) in this column.

### Template categories

For any variable with less than 15 unique values, you will find a list of those values under "value_old". You can modify those values using the "values_new" column. If you prefer to list this information only for variables with much fewer than 15 unique values or with much more than 15 unique values, you have the option of changing this with 

### Template missing value codes

You can find and update missing value codes using the "missings" column of the template. Your template will initially be filled with dots, implying that only values with the default code of NA are currently designated as missing.

You can change this to designate a particular value as missing (e.g., c(-1)), or you can specify a range of values (e.g., c(7,9), >=990).

### Template recodes

The final column in the key template is "recodes." This starts out empty, but you can put in some R code that will automatically transform your data. This is especially useful for dates, which come in so many different sizes and shapes.

## Example

The whas500 is a data set used in Survival Analysis, by Hosmer, Lemeshow, and May. You can download it at 

ftp://ftp.wiley.com/public/sci_tech_med/survival/

along with a description of the variables in the file. When you first read in the data, there are no variable names and no documentation of the various categorical variables.

I was going to add this information the old fashioned way, and then I was inspired to try the kutils program. Here's how it worked.

First you create a key template. You can store this key template as a data frame within R, or you can write it to an Excel file (.xlsx), a comma separated value text file (.csv), or an R serialization data structure (.rds). I chose a csv file.

```{r create-the-template}
fn <- "~/survival-models/data/wiley/whas500.dat"
whas500_raw <- read.table(fn, header=FALSE, as.is=TRUE)
library(kutils)
kn <- "~/survival-models/doc/keys/whas500_key.csv"
whas_key <- keyTemplate(whas500_raw, file=kn)
```

Here's what the csv file looks like.

```{r review-the-template}
cat(readLines(kn), sep="\n")
```

Now you update this file and store it under a different name. For this example, I needed to 

* replace V1 through V22 with descriptive variable names,
* change some of the integer coded variables to factors,
* provide descriptive names for the integer codes, and
* convert the date strings to dates with the as.Date function.

```{r update-the-template}
kn <- "~/survival-models/doc/keys/whas500_updated_key.csv"
cat(readLines(kn), sep="\n")
```

Now apply the updated template.

```{r apply-the-template}
whas_key <- keyImport(kn)
whas500 <- keyApply(whas500_raw, whas_key)
whas500$time_yrs <- whas500$lenfol / 365.25

```

Let's look at the old and new files.

```{r compare-the-old-and-new}
head(whas500_raw)
str(whas500_raw)
head(whas500)
str(whas500)
fn <- "~/survival-models/bin/whas500.RData"
save(whas500, file=fn)
```

## whas100 data set

That was so much fun, let's do it again with the whas100 data set.

```{r whas100-raw}
fn <- "~/survival-models/data/wiley/whas100.dat"
whas100_raw <- read.table(fn, header=FALSE, as.is=TRUE)
library(kutils)
kn <- "~/survival-models/doc/keys/whas100_key.csv"
whas_key <- keyTemplate(whas100_raw, file=kn)
cat(readLines(kn), sep="\n")
```

```{r whas100-updated}
kn <- "~/survival-models/doc/keys/whas100_updated_key.csv"
cat(readLines(kn), sep="\n")
whas_key <- keyImport(kn)
whas100 <- keyApply(whas100_raw, whas_key)
whas100$time_yrs <- whas100$lenfol / 365.25
head(whas100_raw)
str(whas100_raw)
head(whas100)
str(whas100)
fn <- "~/survival-models/bin/whas100.RData"
save(whas100, file=fn)
```

## heroin data set

Here's a data set that used to be found in the DASL website, but appears to have been lost. I have a copy at my website.

```{r heroin-raw}
fn <- "http://www.pmean.com/00files/heroin.txt"
heroin_raw <- read.table(fn, header=FALSE, as.is=TRUE, skip=1)
head(heroin_raw)
heroin1 <- heroin_raw[ , 1:6]
heroin2 <- heroin_raw[ , 7:12]
names(heroin2) <- names(heroin1)
heroin_rejoined <- rbind(heroin1, heroin2)
kn <- "~/survival-models/doc/keys/heroin_key.csv"
heroin_key <- keyTemplate(heroin_rejoined, file=kn)
cat(readLines(kn), sep="\n")
```

```{r heroin-updated}
kn <- "~/survival-models/doc/keys/heroin_updated_key.csv"
cat(readLines(kn), sep="\n")
heroin_key <- keyImport(kn)
heroin <- keyApply(heroin_rejoined, heroin_key)
heroin$time_yrs <- heroin$time/365.25
heroin$i_prison <- as.numeric(heroin$prison_record=="yes")
cut_values <- c(0, 40, 50, 60, 70, 999)
cut_labels <- c("20-40", "45-50", "55-60", "65-70", "75-110")
heroin$dose_group <- cut(
  heroin$methadone_dose, 
  cut_values, 
  cut_labels)

head(heroin)
str(heroin)
fn <- "~/survival-models/bin/heroin.RData"
save(heroin, file=fn)
```

## psychiatric patients

This is a data set of patients admitted to a psychiatric unit.

"Survival data on 26 psychiatric inpatients admitted to the University of Iowa hospitals during the years 1935–1948. This sample is part of a larger study of psychiatric inpatients discussed by Tsuang and Woolson (1977). Data for each patient consists of age at first admission to the hospital, sex, number of years of follow-up (years from admission to death or censoring) and patient status at the follow-up time. Ultimately we want to know if male psychiatric patients and female psychiatric patients are more likely to die than the general public. The data from the hospitals are displayed in the table below (Klein)."

math.usu.edu/jrstevens/biostat/projects2013/pres_LeftTruncation.pdf

```{r psychiatric-patients-raw}
fn <- "~/survival-models/data/psychiatric patients.csv"
psychiatric_raw <- read.csv(fn, header=FALSE, as.is=TRUE, skip=1)
head(psychiatric_raw)
kn <- "~/survival-models/doc/keys/psychiatric_key.csv"
psychiatric_key <- keyTemplate(psychiatric_raw, file=kn)
cat(readLines(kn), sep="\n")
```

```{r psychiatric-patients-updated}
kn <- "~/survival-models/doc/keys/psychiatric_updated_key.csv"
cat(readLines(kn), sep="\n")
psychiatric_key <- keyImport(kn)
psychiatric <- keyApply(psychiatric_raw, psychiatric_key)
psychiatric$age2 <- psychiatric$age+psychiatric$time
head(psychiatric)
str(psychiatric)
fn <- "~/survival-models/bin/psychiatric.RData"
save(psychiatric, file=fn)
```

## leaders data set

This is a data set of 

"(description)"

(source)

```{r leaders-raw}
fn <- "~/survival-models/data/leader.txt"
leader_raw <- read.table(fn, header=FALSE, as.is=TRUE, skip=1)
head(leader_raw)
kn <- "~/survival-models/doc/keys/leader_key.csv"
leader_key <- keyTemplate(leader_raw, file=kn)
cat(readLines(kn), sep="\n")
```

```{r leaders-updated}
kn <- "~/survival-models/doc/keys/leader_updated_key.csv"
cat(readLines(kn), sep="\n")
leader_key <- keyImport(kn)
leader <- keyApply(leader_raw, leader_key)
leader$start <- 1000 + leader$start
head(leader)
str(leader)
fn <- "~/survival-models/bin/leader.RData"
save(leader, file=fn)
```
