---
title: "homework answers 2"
author: "Steve Simon"
date: "April 30, 2018"
output: html_document
---

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

1. Open the WHAS500 data set in the software program of 
your choice.

```{r read-whas500}
fn <- "~/survival-models/bin/whas500.RData" 
load(fn)
whas500$time_yrs <- whas500$lenfol / 365.25
head(whas500)
```

a. Calculate a Cox regression model for males versus females.
Interpret your result.

This calculation will start with some simpler statistics and graphs to place everything in the proper context.

```{r overall, fig.width=4.5, fig.height=2}
table(whas500$fstat)
par(mar=c(2.1, 2.1, 0.6, 0.6))
km_overall <- survfit(Surv(time_yrs, fstat=="Dead")~1, data=whas500)
plot(km_overall, conf.int=FALSE)

table(whas500$gender, whas500$fstat)
km_by_gender <- 
  survfit(Surv(time_yrs, fstat=="Dead")~gender, data=whas500)
plot(km_by_gender)

survdiff(Surv(time_yrs, fstat=="Dead")~gender, data=whas500)
```

Now you are ready for the Cox proportional hazards model.

```{r cox-model-gender}
cox_gender <- coxph(
  Surv(time_yrs, fstat=="Dead")~gender, data=whas500)
summary(cox_gender)
```

You can also export your Cox regression model to a data frame for further manipulation using the broom package. I am not using the augment function because it has issues with Cox regression.

```{r cox-broom}
glance(cox_gender)
tidy(cox_gender)
```

Instead of using augment, you can get estimates for the survival curves and the cumulative hazard using the survfit function.

```{r survfit, fig.width=4.5, fig.height=2}
par(mar=c(2.1, 2.1, 0.6, 0.6))
n <- data.frame(gender=factor(1:2, labels=c("Male", "Female")))
cox_predict <- survfit(cox_gender, newdata=n)
plot(
  cox_predict$time,
  cox_predict$surv[, 2], type="s")
lines(
  cox_predict$time, 
  cox_predict$surv[, 1], 
  type="s", lty="dotted")
plot(
  cox_predict$time,
  cox_predict$cumhaz[, 2], type="s")
lines(
  cox_predict$time, 
  cox_predict$cumhaz[, 1], 
  type="s", lty="dotted")
```

b. Produce age groups <60, 60-69, 70-79, and >=80. 
Calculate a Cox regression model for age groups. Interpret
your results.

```{r km-age-groups, fig.width=4.5, fig.height=2}
par(mar=c(2.1, 2.1, 0.6, 0.6))
age_breaks <- c(0, 59, 69, 79, 99)
age_labels <- c("<60", "60-69", "70-79", ">=80")
whas500$age_group <- cut(whas500$age, age_breaks, age_labels)
table(whas500$age_group, whas500$fstat)
par(mar=c(2.1, 2.1, 0.6, 0.6))
km_by_age_group <- 
  survfit(Surv(time_yrs, fstat=="Dead")~age_group, data=whas500)
plot(km_by_age_group)
survdiff(Surv(time_yrs, fstat=="Dead")~age_group, data=whas500)

cox_age_group <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age_group, data=whas500)
print(cox_age_group)
summary(cox_age_group)
```

c. Treat age as a continuous variable and calculate a Cox
regression model. Interpret your results.

```{r cox-model-age}
cox_age <- 
  coxph(Surv(time_yrs, fstat=="Dead")~age, data=whas500)
summary(cox_age)
```

The hazard ratio, `r round(exp(tidy(cox_age)$estimate), 3)`, is the estimated relative change in the hazard when age increases by one unit. To get the relative change in the hazard when age increases by a decade, raise to the 10th power.

`r round(exp(tidy(cox_age)$estimate),3)`^10 = 
`r round(exp(tidy(cox_age)$estimate)^10,3)` 

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/homework-answers-2.RData")
```