---
title: "homework answers 7"
author: "Steve Simon"
date: "June 14, 2018"
output: html_document
---

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

1. Open the file, diabetes.csv.

```{r load}
library(timereg)
data(diabetes)
head(diabetes)
diabetes_surv <- Surv(diabetes$time, diabetes$status)
```

a. Calculate and graph a Kaplan-Meier curve comparing treated to untreated eyes, ignoring for now the correlations inherent in this data set. Does it appear as if these survival curves differ? If so, do they appear to violate the assumption of proportional hazards?

```{r km, fig.width=4.5, fig.height=2.5}
survfit(diabetes_surv~treat, data=diabetes) %>%
  tidy                                      %>%
  ggplot(aes(time, estimate, color=strata))  +
  expand_limits(y=0)                         +
  geom_step()
```

b. Calculate and interpret a Cox regression model using treat as an independent variable and id as a cluster effect.

```{r cluster}
coxph(
  diabetes_surv~treat+agedx+cluster(id),
  data=diabetes)
```

c. Calculate and interpret a Cox regression model using treat as an independent variable and id as a frailty effect.

```{r frailty}
coxph(
  diabetes_surv~treat+agedx+frailty(id),
  data=diabetes)
```


```{r save-everything}
save.image("~/survival-models/bin/homework-answers-7.RData")
```