---
title: "homework answers 6"
author: "Steve Simon"
date: "May 27, 2018"
output: html_document
---

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

## The Kaplan-Meier plot and proportional hazards.

```{r read}
fn <- "~/survival-models/bin/whas500.RData"
load(fn)
whas500$time_yrs <- whas500$lenfol / 365.25
```

```{r km-plots, fig.width=4.5, fig.height=2.5}
whas500 %$% Surv(time_yrs, fstat=="Dead") -> whas500_surv
hr_cuts <- c(0, 70, 85, 100, 999)
whas500                                    %>%
  mutate(hr_categories=cut(hr, hr_cuts))   %$%
  survfit(whas500_surv~hr_categories)      %>%
  tidy                                      -> km_hr
km_hr                                       %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()

diasbp_cuts <- c(0, 60, 70, 80, 999)
whas500                                     %>%
  mutate(
    diasbp_categories=
      cut(diasbp, diasbp_cuts))             %$%
  survfit(whas500_surv~diasbp_categories)   %>%
  tidy                                      -> km_diasbp
km_diasbp                                   %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()

whas500                                     %$%
  survfit(whas500_surv~chf)                 %>%
  tidy                                      -> km_chf
km_chf                                      %>%
  ggplot(aes(time, estimate, color=strata))  +
    geom_step()
```

## Complementary log-log plot

```{r log-log, fig.width=4.5, fig.height=2.5}
km_hr                                 %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()

km_diasbp                             %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()

km_chf                                %>%
  mutate(cll=log(-log(estimate)))     %>%
  ggplot(aes(time, cll, color=strata)) +
    geom_step()

```

## Schoenfeld residuals.

```{r schoenfeld, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 4.1, 0.6, 0.6))
cox_mv <- coxph(
  Surv(time_yrs, fstat=="Dead") ~
    hr +
    diasbp +
    chf,
  data=whas500)
cox_schoenfeld <- cox.zph(cox_mv)
print(cox_schoenfeld)
plot(cox_schoenfeld)

```

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/homework-answers-6.RData")
```