---
title: "supplement 8"
author: "Steve Simon"
date: "June 3, 2018"
output: html_document
---

This file creates various formulas and graphs needed to illustrate the underlying theory behind the data analyses in survival-lecture-8.Rmd. I am sharing this for those who are curious, but you are not responsible for learning or using the code shown in this supplement.

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

```{r set-parameters}
lambda <- 1.08
```

Suppose you have a population of women with ages that are normally distributed with a mean of 50 and a standard deviation of 10. Also assume that the hazard increases by  factor of `r lambda` for each year of age. This is what the population of survival functions would look like.

```{r population, fig.width=4.5, fig.height=2.5}
n <- 99
m <- 301
p <- (1:n)/(n+1)
a <- qnorm(p, 50, 10)
# a <- c(a, 0)
ax <- rep(a, each=m)
al <- rep(0.15, length(ax))
co <- ifelse(ax==50, "red", "black")
al <- ifelse(ax==50, 1, 0.15)
# co[ax==0] <- "green"
# al[ax==0] <- 1
t <- seq(0, 3, length=m)
data.frame(id=rep(1:n, each=m)) %>%
  mutate(age=rep(a, each=m))    %>%
  mutate(time=rep(t, n))        %>%
  mutate(hr=lambda^(age-50))    %>%
  mutate(s=exp(-t*hr))           -> df

df %>%
  group_by(time) %>%
  summarize(s=mean(s)) %>%
  ungroup -> dm

df %>%
  ggplot(aes(time, s, group=id)) +
    geom_line(alpha=al, color=co) -> d0
d0

d0 + stat_summary(fun.y="mean", color="green", group=1, geom="line") +
   stat_summary(fun.y="median", color="blue", group=1, geom="line") 
```

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/supplement-8.RData")
```