---
title: "supplement 8"
author: "Steve Simon"
date: "June 3, 2018"
output: html_document
---

This file creates various formulas and graphs needed to illustrate the underlying theory behind the data analyses in survival-lecture-8.Rmd. I am sharing this for those who are curious, but you are not responsible for learning or using the code shown in this supplement.

```{r render, eval=FALSE, echo=FALSE}
# run the commands in this section to store the output
# in the results folder.
library(rmarkdown)
f <- "~/survival-models/src/supplement-8.Rmd"
o <- "~/survival-models/results"
render(f, output_dir=o)
```

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

## competing risks

The leader data set is described in Chapter 6 of Alison. You can get the data, and the SAS code at

http://ftp.sas.com/samples/A55233

The prostate survival data set is available in the asaur package.

```{r prostate}
head(prostateSurvival)
```

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/supplement-8.RData")
```