---
title: "survival lecture 7"
author: "Steve Simon"
date: "April 8, 2018"
output: html_document
---

Lecture 7. Time varying covariates in a Cox model. Time varying covariates allow you to account for non-proportional hazards and can model settings where patients switch from one therapy to another. You will code data for time-varying covariates, fit time-varying models, and interpret the results.

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

## Time-varying covariates

Page 221 of Hosmer, Lemeshow, and May describes the grace data set, which you can use for time-varying covariates.

```{r read-grace}
fn <- "~/survival-models/data/wiley/GRACE1000.dat" 
grace <- read.table(fn, header = FALSE) 
names(grace) <- c(
  "id",
  "days",
  "death",
  "revasc",
  "revascdays",
  "los",
  "age",
  "sysbp",
  "stchange")
head(grace)
```

## Also for discussion

* Stratified analysis. Stratify whas500 by year=1, 2, 3. This would use the actg320 data. Compare output to Table 7.1.

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/survival-lecture-7.RData")
```