---
title: "survival lecture 7"
author: "Steve Simon"
date: "April 8, 2018"
output: html_document
---

Lecture 7. Time varying covariates in a Cox model. Time varying covariates allow you to account for non-proportional hazards and can model settings where patients switch from one therapy to another. You will code data for time-varying covariates, fit time-varying models, and interpret the results.

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

Lecture 3. Planning and data management issues for survival data. Planning a study with a survival outcome requires you to specify both the number of patients and the duration of follow-up time. You’ll compute power for hypothetical studies and compare power across different research designs. Then you’ll review the data management needs of a survival study, with a special emphasis on the problems associated with date variables.

## Cox multivariate regression

Recall that we ran a log rank test for gender using the WHAS100 data file. Let's calculate a Cox regression model for the same data and compare it to the log rank test.

```{r read-whas100}
fn <- "~/survival-models/data/wiley/whas100.dat" 
whas100 <- read.table(fn, header = FALSE) 
names(whas100) <- 
  c("id", "admitdate", "foldate",
    "los", "lenfol", "fstat",
    "age", "gender", "bmi")
whas100$time_yrs <- whas100$lenfol / 365.25
head(whas100)
whas100_surv <- Surv(whas100$time_yrs, whas100$fstat)
```

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/survival-lecture-7.RData")
```