---
title: "survival lecture 8"
author: "Steve Simon"
date: "May 27, 2018"
output: html_document
---

Lecture 8. Competing risk models.

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

Read in the leader data set.

```{r read-leader}
fn <- "~/survival-models/bin/leader.RData"
load(fn)
leader <- leader[is.finite(leader$age), ]
```

Calculate some simple descriptive statistics.

```{r descriptives}
summary(leader$years)
table(leader$lost)
table(leader$manner)
summary(leader$start)
table(leader$military)
summary(leader$age)
table(leader$conflict)
summary(leader$loginc)
summary(leader$growth)
summary(leader$pop)
summary(leader$land)
summary(leader$literacy)
table(leader$region)
```

First, let's look at this model in a binary fashion, with lost = "still in power" as censored observations and "constitutional exit" and "natural death" and "non-constitutional exist" as events.

```{r binary-survival}
leader$event <- 
  (leader$lost=="constitutional exit") +
  (leader$lost=="natural death") +
  (leader$lost=="nonconstitutional exit")

table(leader$lost, leader$event, useNA="ifany")
```

Always check your counts to make sure you have enough events in each group.

```{r binary-counts}
table(leader$manner, leader$event)
table(leader$military, leader$event)
table(leader$conflict, leader$event)
table(leader$region, leader$event)
```

Create cut-points for continuous outcomes.

```{r cutpoints}
leader$start_c <- 
  cut(
    leader$start,
    c(1959, 1969, 1979, 1987),
    c("60s", "70s","80s"  )
)
table(leader$start_c)
leader$age_c <- 
  cut(
    leader$age,
    c(0, 39, 59, 89),
    c("0-39", "40-59", "60+")
)
table(leader$age_c)
leader$loginc_c <- 
  cut(
    leader$loginc,
    c(log(c(1, 200, 500, 100000))), 
    c("$0-200", "$200-500", "$500+")
)
table(leader$loginc_c)
leader$growth_c <- 
  cut(
    leader$growth,
    c(-10, 0, 3.9, 10),
    c("neg", "0-3.9%", "4%+")
)
table(leader$growth_c)
leader$pop_c <-
  cut(
    leader$pop,
    c(0, 1, 10, 10000),
    c("under 1m", "1-10m", "10m+")
)
table(leader$pop_c)
leader$land_c <- 
  cut(
    leader$land, 
    c(0, 100, 1000, 10000),
    c(
      "under 100 sq km",
      "100-1,000 sq km",
      "1,000-10,000 sq km")
)
table(leader$land_c)
leader$literacy_c <- 
  cut(
    leader$literacy,
    c(0, 50, 75, 100),
    c("0-50%", "50-75%", "75-100%")
)
table(leader$literacy_c)
```

```{r binary-km, fig.width=4.5, fig.height=2.5}
binary_surv <- Surv(leader$years, leader$event)

draw_km <- function(v) {
  survfit(binary_surv~leader[ , v], data=leader)   %>%
    tidy                                           %>%
    ggplot(aes(time, estimate, color=strata))  +
      ylab(v)                                       +
      expand_limits(y=0:1)                          +
      geom_step()                                  %>%
  return  
}

draw_km("manner")
draw_km("start_c")
draw_km("military")
draw_km("age_c")
draw_km("conflict")
draw_km("loginc_c")
draw_km("growth_c")
draw_km("pop_c")
draw_km("land_c")
draw_km("literacy_c")
draw_km("region")
```


Fit univariate Cox models

```{r univariate-cox}
coxph(binary_surv~manner, data=leader)
coxph(binary_surv~start, data=leader)
coxph(binary_surv~military, data=leader)
coxph(binary_surv~age, data=leader)
coxph(binary_surv~conflict, data=leader)
coxph(binary_surv~loginc, data=leader)
coxph(binary_surv~growth, data=leader)
coxph(binary_surv~pop, data=leader)
coxph(binary_surv~land, data=leader)
coxph(binary_surv~literacy, data=leader)
coxph(binary_surv~region, data=leader)
```

Fit a multivariate model

```{r mv}
binary_mv <-
  coxph(
    binary_surv~
      manner + 
      military +
      age,
    data=leader
  )
summary(binary_mv)
```

Compare the Martingale residuals to each independent variable already in the model to look for non-linearities.

```{r martingale1, fig.width=4.5, fig.height=2.5}
leader$r <- residuals(binary_mv)
ggplot(leader, aes(manner, r))   +
  geom_boxplot()

ggplot(leader, aes(military, r)) +
  geom_boxplot()                 +
  geom_smooth(se=TRUE)

ggplot(leader, aes(age, r))      +
  geom_point()                   +
  geom_smooth(se=TRUE)
```

Compare the Martingale residuals to variables not yet in the model.

```{r martingale2, fig.width=4.5, fig.height=2.5}
leader$r <- residuals(binary_mv)
ggplot(leader, aes(start, r))    +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(conflict, r)) +
  geom_boxplot()

ggplot(leader, aes(loginc, r))   +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(growth, r))   +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(pop, r))      +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(log(pop), r)) +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(log(land), r))     +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(literacy, r)) +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(region, r))   +
  geom_boxplot()
```

Update your multivariate model

```{r mv2}
binary_mv2 <-
  coxph(
    binary_surv~
      manner + 
      military +
      age +
      loginc +
      region,
    data=leader
  )
summary(binary_mv2)
```

Computet the Schoenfeld residuals to look for non-proportional hazards.

```{r schoenfeld, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
leader_residuals <- cox.zph(binary_mv2)
plot(leader_residuals)
```

## Competing risk models

```{r simple-km-curves, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
surv_object <- 
  Surv(
    leader$years, 
    leader$lost,
    type="mstate")

km_overall <- survfit(surv_object~1)
plot(
  km_overall$time, 
  km_overall$pstate[ , 1], 
  ylim=0:1, type="s")
plot(
  km_overall$time,
  km_overall$pstate[ , 2], 
  ylim=0:1, type="s")
plot(
  km_overall$time,
  km_overall$pstate[ , 3],
  ylim=0:1, type="s")
plot(
  km_overall$time,
  km_overall$pstate[ , 4],
  ylim=0:1, type="s")

km_single <-
  survfit(
    Surv(
      leader$years, 
      leader$lost!="still in power")~1)
plot(km_single, ylim=0:1, conf.int=FALSE)
points(
  km_overall$time, 
  km_overall$pstate[ , 4], 
  col="red")
head(tidy(km_overall))
ggplot(
  data=tidy(km_overall),
  aes(time, estimate, fill=state)) + 
  geom_col(width=1)

km_manner <- survfit(surv_object~leader$manner)
str(km_manner)
km_tidy <- tidy(km_manner)
head(km_tidy)
categorical_list <- c(
  "manner",
  "military",
  "conflict",
  "region"
)
continuous_list <- c(
  "start",
  "loginc",
  "growth",
  "pop",
  "land",
  "literacy"
)
```

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/survival-lecture-8.RData")
```