---
title: "survival lecture 1 logrank details"
author: "Steve Simon"
date: "March 24, 2018"
output: html_document
---

In my Powerpoint lecture, I had a few problems with notation and produced some confusion on how the logrank test actually works. Let me try again with this program. 

This file does not need any special libraries other than the ones listed below.

```{r load-libraries, echo=FALSE}
library(knitr)
library(survival)
```

Consider a small data set with the following survival times:

Male 6, 44+, 98, 114
Female 14, 44, 89+, 98, 104

To compare this properly, you need to create a table with the number of deaths, the number of censored values, and the number at risk, both overall and for the two groups.

First list the times in order. You can exclude times with only censored values, if you like.

```{r table1, echo=FALSE}
ti <- c(6, 14, 44, 89, 98, 104, 114)
```

`r kable(data.frame())`

Count the deaths in each group . d1i is deaths among the women, and d0i is deaths among the men.

```{r table2, echo=FALSE}
d1i <- c(0, 1, 1, 0, 1, 1, 0)
d0i <- c(1, 0, 0, 0, 1, 0, 1)
```

`r kable(data.frame(ti, d1i, d0i))`

Count the censored values in each group. This is not really necessary for computation, but it makes it a bit easier to follow the results.

```{r table3, echo=FALSE}
c1i <- c(0, 0, 0, 1, 0, 0, 0)
c0i <- c(0, 0, 1, 0, 0, 0, 0)
```


Calculate the number at risk in each group.

```{r table4, echo=FALSE}
n1i <- c(5, 5, 4, 3, 2, 1, 0)
n0i <- c(4, 3, 3, 2, 2, 1, 1)
```

`r kable(data.frame(ti, d1i, c1i, n1i, d0i, c0i, n0i))`

Now calculate totals for both groups. You can drop group 0 and the censored values. Also drop the time point 89, which had only one censored value and no deaths.

```{r table5, echo=FALSE}
di <- d1i+d0i
ci <- c1i+c0i
ni <- n1i+n0i
df <- data.frame(ti, d1i, n1i, di, ni)[ -4, ]

```

`r kable(df)`

So far, the only thing you have done is count, subtract, and add. Here's where it gets a bit tricky.

If the two groups had equal survival probabilities, then you can estimate the conditional probability of death at $t_i$ as $p_i=\frac{d_i}{n_i}$. The expected number of deaths in the females is $n_{1i} p_i$. Calculate the observed number of deaths among the females minus the expected number of deaths.

```{r table6, echo=FALSE}
df$pi <- df$di/df$ni
df$e1i <- round(df$n1i*df$pi, 3)
df$o_minus_e <- df$d1i-df$e1i
```

`r kable(df)`

Notice that there is a mix of positive and negative numbers for the observed minus expected, but slightly more positive values. The total observed minus expected is `r sum(df$o_minus_e)`.

If the death count were binomial, the variance at each time point would be $n_{1i} p_i (1-p_i)$ but you also have to adjust for the small sample sizes with a finite population correction factor, $\frac{n_i-n_{1i}}{n_i-1}$. 

```{r table7, echo=FALSE}
df$vi <- round(df$n1i*df$pi*(1-df$pi)*(df$ni-df$n1i)/(df$ni-1), 3)
df$vi[6] <- 0
```

`r kable(df)`

The sum of the variances is `r sum(df$vi)` and you compare that to the square of the sum of observed minus expected `r sum(df$o_minus_e)^2`.

Save everything for possible re-use.

```{r save-everything, echo=FALSE}
save.image("~/survival-models/bin/survival-lecture-1-logrank-details.RData")
```