---
title: "survival lecture 8"
author: "Steve Simon"
date: "May 27, 2018"
output: html_document
---

Lecture 8. Competing risk models.

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

Read in the leader data set.

```{r read-leader}
fn <- "~/survival-models/bin/leader.RData"
load(fn)
leader <- leader[is.finite(leader$age), ]
```

Calculate some simple descriptive statistics.

```{r descriptives}
summary(leader$years)
table(leader$lost)
table(leader$manner)
summary(leader$start)
table(leader$military)
summary(leader$age)
table(leader$conflict)
summary(leader$loginc)
summary(leader$growth)
summary(leader$pop)
summary(leader$land)
summary(leader$literacy)
table(leader$region)
```

First, let's look at this model in a binary fashion, with lost = "still in power" as censored observations and "constitutional exit" and "natural death" and "non-constitutional exit" as events.

```{r binary-survival}
leader$event <- 
  (leader$lost=="constitutional exit") +
  (leader$lost=="natural death") +
  (leader$lost=="nonconstitutional exit")

table(leader$lost, leader$event, useNA="ifany")
```

Always check your counts to make sure you have enough events in each group.

```{r binary-counts}
table(leader$manner, leader$event)
table(leader$military, leader$event)
table(leader$conflict, leader$event)
table(leader$region, leader$event)
```

Create cut-points for continuous outcomes.

```{r cutpoints}
leader$start_c <- 
  cut(
    leader$start,
    c(1959, 1969, 1979, 1987),
    c("60s", "70s","80s"  )
)
table(leader$start_c)
leader$age_c <- 
  cut(
    leader$age,
    c(0, 39, 59, 89),
    c("0-39", "40-59", "60+")
)
table(leader$age_c)
leader$loginc_c <- 
  cut(
    leader$loginc,
    c(log(c(1, 200, 500, 100000))), 
    c("$0-200", "$200-500", "$500+")
)
table(leader$loginc_c)
leader$growth_c <- 
  cut(
    leader$growth,
    c(-10, 0, 3.9, 10),
    c("neg", "0-3.9%", "4%+")
)
table(leader$growth_c)
leader$pop_c <-
  cut(
    leader$pop,
    c(0, 1, 10, 10000),
    c("under 1m", "1-10m", "10m+")
)
table(leader$pop_c)
leader$land_c <- 
  cut(
    leader$land, 
    c(0, 100, 1000, 10000),
    c(
      "under 100 sq km",
      "100-1,000 sq km",
      "1,000-10,000 sq km")
)
table(leader$land_c)
leader$literacy_c <- 
  cut(
    leader$literacy,
    c(0, 50, 75, 100),
    c("0-50%", "50-75%", "75-100%")
)
table(leader$literacy_c)
```

Take a look at the calculations involved.

```{r calculations-km}
leader_tally <- table(leader$years, leader$lost)
leader_tally
fn <- "~/survival-models/data/leader_tally.csv"
write.csv(leader_tally, file=fn, row.names=FALSE)
```

```{r binary-km, fig.width=4.5, fig.height=2.5}
binary_surv <- Surv(leader$years, leader$event)

draw_km <- function(v) {
  survfit(binary_surv~leader[ , v], data=leader)   %>%
    tidy                                           %>%
    ggplot(aes(time, estimate, color=strata))  +
      ylab(v)                                       +
      expand_limits(y=0:1)                          +
      geom_step()                                  %>%
  return  
}

draw_km("manner")
draw_km("start_c")
draw_km("military")
draw_km("age_c")
draw_km("conflict")
draw_km("loginc_c")
draw_km("growth_c")
draw_km("pop_c")
draw_km("land_c")
draw_km("literacy_c")
draw_km("region")
```


Fit univariate Cox models

```{r univariate-cox}
coxph(binary_surv~manner, data=leader)
coxph(binary_surv~start, data=leader)
coxph(binary_surv~military, data=leader)
coxph(binary_surv~age, data=leader)
coxph(binary_surv~conflict, data=leader)
coxph(binary_surv~loginc, data=leader)
coxph(binary_surv~growth, data=leader)
coxph(binary_surv~pop, data=leader)
coxph(binary_surv~land, data=leader)
coxph(binary_surv~literacy, data=leader)
coxph(binary_surv~region, data=leader)
```

Fit a multivariate model

```{r mv}
binary_mv <-
  coxph(
    binary_surv~
      manner + 
      military +
      age,
    data=leader
  )
summary(binary_mv)
```

Compare the Martingale residuals to each independent variable already in the model to look for non-linearities.

```{r martingale1, fig.width=4.5, fig.height=2.5}
leader$r <- residuals(binary_mv)
ggplot(leader, aes(manner, r))   +
  geom_boxplot()

ggplot(leader, aes(military, r)) +
  geom_boxplot()                 +
  geom_smooth(se=TRUE)

ggplot(leader, aes(age, r))      +
  geom_point()                   +
  geom_smooth(se=TRUE)
```

Compare the Martingale residuals to variables not yet in the model.

```{r martingale2, fig.width=4.5, fig.height=2.5}
leader$r <- residuals(binary_mv)
ggplot(leader, aes(start, r))    +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(conflict, r)) +
  geom_boxplot()

ggplot(leader, aes(loginc, r))   +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(growth, r))   +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(pop, r))      +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(log(pop), r)) +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(log(land), r))     +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(literacy, r)) +
  geom_point()                   +
  geom_smooth(se=TRUE)

ggplot(leader, aes(region, r))   +
  geom_boxplot()
```

Update your multivariate model

```{r mv2}
binary_mv2 <-
  coxph(
    binary_surv~
      manner + 
      military +
      age +
      loginc +
      region,
    data=leader
  )
summary(binary_mv2)
```

Computet the Schoenfeld residuals to look for non-proportional hazards.

```{r schoenfeld, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
leader_residuals <- cox.zph(binary_mv2)
plot(leader_residuals)
```

## Competing risk models

If there are two or more causes of death and you are interested in developing a model for the individual causes, then you need a competing risk model.

The competing risk model assumes that the overal hazard function, h(t), can be written as

$h(t) = \sum_j h_j(t)$,

where $h_j(t)$ is the hazard function for the jth cause of death.

This implies that the overall cumulative hazard, H(t), is equal to

$H(t) = \sum_j H_j(t)$

where $H_j(t)$ is the cumulative hazard for the jth cause of death.

With a bit of math, you can easily show that the overall cumulative density function, F(t)=1-S(t), can be decomposed into

$F(t) = \sum_j F_j(t)$

where 

$F_j(t) = \int_0^t f_j(u) S(u)dt$ 

Notice that $F_j$ is not a cumulative distribution function. In fact, as t approaches infinity, $F_j(t)$ approaches $P_j$, the probability for the jth cause of death. $F_j$ is called the subdistribution or the cumulative incidence function.

You can estimate $F_j$ with the formula

$\hat{F}_j(t) = \sum_{t_{(i)}<t} \hat{h}_j(t_{(i)}) \hat{S}(t_{(i-1)})$

where $\hat{S}$ is the overall Kaplan-Meier estimate of survival, 

$\hat{h}_j(t_{(i)}) = \frac{d_{ij}}{n_i}$,

and $d_{ij}$ and $n_j$ are the number of deaths at time $t_{(i)}$ due to cause j and the number of patients at risk at time $t_{(i)}$.

```{r simple-km-curves, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
surv_object <- 
  Surv(
    leader$years, 
    leader$lost,
    type="mstate")

km_overall <- survfit(surv_object~1)
plot(
  km_overall$time, 
  km_overall$pstate[ , 1], 
  ylim=0:1, type="s")
plot(
  km_overall$time,
  km_overall$pstate[ , 2], 
  ylim=0:1, type="s")
plot(
  km_overall$time,
  km_overall$pstate[ , 3],
  ylim=0:1, type="s")
plot(
  km_overall$time,
  km_overall$pstate[ , 4],
  ylim=0:1, type="s")
```

```{r show-equivalence, fig.width=4.5, fig.height=2.5}
km_single <-
  survfit(
    Surv(
      leader$years, 
      leader$lost!="still in power")~1)
plot(km_single, ylim=0:1, conf.int=FALSE)
points(
  km_overall$time, 
  km_overall$pstate[ , 4], 
  col="red")
head(tidy(km_overall))
km_overall %>% tidy -> km_overall_tidy
km_overall_tidy %>%
  filter(time==24) %>%
  mutate(time=25) %>%
  bind_rows(km_overall_tidy) -> km_overall_tidy
ggplot(
  data=km_overall_tidy,
  aes(time, estimate, fill=state)) + 
  scale_y_reverse()                +
  expand_limits(y=1)               +
  geom_col(width=1)
```

Let's look at a couple of subsets.

```{r subsets, fig.width=4.5, fig.height=2.5}
sb1 <- leader$manner=="constitutional ascent"
km_sb1 <- survfit(surv_object~1, subset=sb1)
km_sb1 %>% tidy -> km_sb1_tidy
km_sb1_tidy %>%
  filter(time==20) %>%
  mutate(time=21) %>%
  bind_rows(km_sb1_tidy) -> km_sb1_tidy
km_sb1_tidy %>%
  filter(time==24) %>%
  mutate(time=25) %>%
  bind_rows(km_sb1_tidy) -> km_sb1_tidy
ggplot(
  data=km_sb1_tidy, 
  aes(time, estimate, fill=state)) +
  scale_y_reverse()                +
  expand_limits(y=1)               +
  expand_limits(x=27)              +
  geom_col(width=1)
sb2 <- leader$manner=="nonconstitutional ascent"
km_sb2 <- survfit(surv_object~1, subset=sb2)
km_sb2 %>% tidy -> km_sb2_tidy
for (i in 22:27) {
  km_sb2_tidy %>%
    filter(time==21) %>%
    mutate(time=i) %>%
    bind_rows(km_sb2_tidy) -> km_sb2_tidy
}
ggplot(
  data=km_sb2_tidy, 
  aes(time, estimate, fill=state)) +
  scale_y_reverse()                +
  expand_limits(y=1)               +
  expand_limits(x=27)              +
  geom_col(width=1)
```


```{r average}
n <- 1000
y1 <- rexp(n, 1)
y2 <- rexp(n, 2)
y3 <- rexp(n, rep(1:2, n/2))
y4 <- rexp(n, 1.5)
cens <- rep(1, n)
km1 <- tidy(survfit(Surv(y1, cens)~1))
km2 <- tidy(survfit(Surv(y2, cens)~1))
km3 <- tidy(survfit(Surv(y3, cens)~1))
km4 <- tidy(survfit(Surv(y4, cens)~1))
plot(km1$time, km1$estimate, type="l", col="red")
lines(km2$time, km2$estimate, col="blue")
lines(km3$time, km3$estimate, col="purple")
lines(km4$time, km4$estimate, col="purple", lty="dotted")
```

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/survival-lecture-8.RData")
```