---
title: "survival lecture 5"
author: "Steve Simon"
date: "May 27, 2018"
output: html_document
---

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

## Exponential model

The exponential model is the simplest parametric model. The probability density function for the "standard" exponential distribution is

$f(t)=e^{-t}$

and the survival curve is

$S(t)=e^{-t}$.

The plot of the survival curve is 

```{r density-1, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
t <- seq(0, 3, length=1000)
plot(t, exp(-t), type="l", ylim=c(0,1))
```

You can include a rate parameter, $\lambda$, to create a density,

$f(t; \lambda)=\lambda e^{-\lambda t}$

and the survival curve is

$S(t; \lambda)=e^{-\lambda t}$

There is an equivalent form using a scale parameter, $\theta$,

$f(t; \theta) = \frac{1}{\theta} e^{-t/\theta}$.

$S(t; \theta) = e^{-t/\theta}$

This is the density function for an exponential distribution with  $\theta=2$.

```{r density-2, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
theta <- 2
plot(t, exp(-t/theta), type="l", ylim=c(0,1))
lines(t, exp(-t), lty="dotted")
```

If theta is larger than 1, the effect is to stretch out the exponential and slow down time.

This is the density function for an exponential distribution with  $\theta=0.5$.

```{r density-0.5, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
theta <- 0.5
plot(t, exp(-t/theta), type="l", ylim=c(0,1))
lines(t, exp(-t), lty="dotted")
```

The accelerated time model replaces $\theta$ with

$e^{(\beta_0+\beta_1 X)}$

This produces the survival curve

$S(t, X, \beta_0, \beta_1)=e^{-t/e^{(\beta_0+\beta_1 X)}}$

The values of $\beta_0$ and $\beta_1$ will end up stretching or shrinking the time scale.

The pth percentile of the accelerated time model is

$-ln(1-p) e^{\beta_0+\beta_1 X}$ 

and the ratio of two percentiles, one with X=$X_1$ and the other with X=$X_2$ is

$\frac{-ln(1-p) e^{\beta_0+\beta_1 X_1}} {-ln(1-p) e^{\beta_0+\beta_1 X_2}} = e^{\beta_1 (X_1-X_2)}$

If $X_1$ is one unit larger than $X_2$, this reduces to $e^{\beta_1}$.

This may look quite different than the model we used for Kaplan-Meier curves and the Cox proportional hazards model, but it actually is not. The hazard function is 

$h(t, x, \beta_0, \beta_1)=e^{-(\beta_0+\beta_1 x)}$.

Notice that the hazard is constant with respect to t. The baseline hazard, the hazard when X=0 is

$h_0(t)=e^{-\beta_0}$

The hazard ratio for a subject with $x=x_1$ compared to a subject with $x=x_2$ is

$e^{-\beta_1 (x_1-x_2)}$

## Weibull model

The standard Weibull distribution has a density

$f(t, k)=k t^{k-1} e^{-t^k}$

and a survival function

$S(t, k)=e^{-t^\gamma}$.

You can add a scale parameter, $\theta$, to get 

$f(t, k, \theta) = \frac {k} {\theta} \left( \frac {t} {\theta} \right)^{k-1} e^{-\left( \frac {t} {\theta} \right)^k}$

and a survival function

$S(t, k, \theta)=e^{-\left(\frac{t}{\theta}\right)^k}$.

The value of k controls whether the Weibull distribution has an increasing hazard (k>1), constant hazard (k=1), or decreasing hazard (k<1).

```{r standard-weibulls, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
plot(t, 1-pweibull(t, shape=2), type="l", ylim=c(0,1))
lines(t, 1-pweibull(t, shape=1), lty="dotted")
plot(t, 1-pweibull(t, shape=0.5), type="l", ylim=c(0,1))
lines(t, 1-pweibull(t, shape=1), lty="dotted")
```

The scale parameter, $\theta$, will stretch or squeeze the survival curve along the time axis.

```{r scaled-weibulls, fig.width=4.5, fig.height=2.5}
par(mar=c(2.6, 2.6, 0.6, 0.6))
plot(t, 1-pweibull(t, shape=2, scale=2), type="l", ylim=c(0,1))
lines(t, 1-pweibull(t, shape=2, scale=1), lty="dotted")
plot(t, 1-pweibull(t, shape=2, scale=0.5), type="l", ylim=c(0,1))
lines(t, 1-pweibull(t, shape=2, scale=1), lty="dotted")
```

You can fit an accelerated time model for the Weibull distribution as well.

$S(t, k, X, \beta_0, \beta_1) = e^{-\left( \frac {t} {e^{\beta_0 + \beta_1 x}} \right)^k}$.


```{r residuals}
r_response <- residuals(weibull_model, type="response")
r_deviance <- residuals(weibull_model, type="deviance")
r_dfbeta <- residuals(weibull_model, type="dfbeta")
r_dfbetas <- residuals(weibull_model, type="dfbetas")
r_working <- residuals(weibull_model, type="working")
r_ldcase <- residuals(weibull_model, type="ldcase")
r_ldresp <- residuals(weibull_model, type="ldresp")

head(r_response)
head(r_deviance)
head(r_dfbeta)
head(r_dfbetas)
head(r_working)
head(r_ldcase)
head(r_ldresp)

p_response <- predict(weibull_model, type="response")
p_linear <- predict(weibull_model, type="linear")
p_terms <- predict(weibull_model, type="terms")

head(p_response)
head(p_linear)
exp(head(p_linear))
head(p_terms)

term_totals <- apply(p_terms, 1, sum)
head(term_totals)
```

The hazard function for the Weibull distribution is

$h(t, k) = k t^{k-1}$

There is an alternate formulation for the Weibull regression model.

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/survival-lecture-5-supplement.RData")
```