---
title: "homework answers 5"
author: "Steve Simon"
date: "May 27, 2018"
output: html_document
---

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

Open the WHAS500 data set in the software program of 
your choice.

```{r open}
fn <- "~/survival-models/bin/whas500.RData"
load(fn)
head(whas500)
```

Calculate and graph on the same graph an overall Kaplan-Meier survival curve, a survival curve for an exponential model with no independent variables, and a survival curve for a Weibull model with no independent variables.

```{r graph}
whas500_surv <- Surv(whas500$time_yrs, whas500$fstat=="Dead")
km_overall <- survfit(whas500_surv~1)

exp_null_model <- survreg(whas500_surv~1, dist="exponential")
pct <- (1:999)/1000
exp_null_model %>%
  predict(
    newdata=data.frame(one=1),
    type="quantile",
    p=pct)                     %>%
  data.frame                   %>%
  set_names("time")            %>%
  mutate(estimate=rev(pct))    %>%
  mutate(grp="Exponential")        %>%
  filter(time<=10)             -> exp_survival
weibull_null_model <- survreg(whas500_surv~1, dist="weibull")
weibull_null_model %>%
  predict(
    newdata=data.frame(one=1),
    type="quantile",
    p=pct)                     %>%
  data.frame                   %>%
  set_names("time")            %>%
  mutate(estimate=rev(pct))    %>%
  mutate(grp="Weibull")        %>%
  filter(time<=10)             -> weibull_survival

km_overall                     %>%
  tidy                         %>%
  mutate(grp="Kaplan-Meier")   %>%
  bind_rows(exp_survival)      %>%
  bind_rows(weibull_survival)  -> combined_survivals

combined_survivals             %>%
  ggplot(aes(time, estimate))   +
    expand_limits(y=0)          +
    geom_step(aes(color=grp))

```

Does there appear to be a difference between the exponential model and the Weibull model? Does either model match the Kaplan-Meier curve?

Answer: The exponential model is a poor fit. It overestimates survival early and underestimates it late. The Weibull model is not perfect, but it matches much more closely.

Calculate a Weibull regression model with bmi, age, gender, and a age by gender interaction. Estimate the impact of a 5 unit change in BMI on the survival percentiles, holding all of the other variables constant.

```{r fit-weibull}
weibull_full_model <- survreg(
  whas500_surv~bmi + age*gender, 
  data=whas500,
  dist="weibull")
summary(weibull_full_model)
weibull_full_model          %>%
  tidy                      %>%
  filter(term=="bmi")       %>%
  select(estimate)          %>%
  mutate(hr=exp(estimate))  %>%
  mutate(hr5=hr^5)
```

Calculate a confidence interval for the scale parameter of the Weibull model.

```{r scale-ci}
weibull_full_model                           %>%
  tidy                                       %>%
  filter(term=="Log(scale)")                 %>%
  select(estimate, std.error)                %>%
  mutate(scale=exp(estimate))                %>%
  mutate(lo=exp(estimate-1.96*std.error))    %>%
  mutate(hi=exp(estimate+1.96*std.error))
```

Interpret this interval.

Answer: Since this interval excludes the value of 1, you can conclude that the exponential model is not a good fit for the data.

Save everything for possible later re-use.

```{r save-everything}
save.image("~/survival-models/bin/homework-answers-5.RData")
```