---
title: "homework answers 8"
author: "Steve Simon"
date: "June 18, 2018"
output: html_document
---

This file does not need any special libraries other than the ones listed below. Many of the data sets in this program use data from Hosmer, Lemeshow, and May. I made one minor change, however, which was to force all the variable names to lower case.

```{r load-libraries}
library(broom)
library(dplyr)
library(ggplot2)
library(magrittr)
library(survival)
library(tidyr)
```

~/survival-models/doc/homework-8.txt

1. Open the file, leader.txt

Calculate a competing risk model comparing Latin America to the other three regions. What are the major differences?

```{r read}
fn <- "~/survival-models/bin/leader.RData"
load(fn)
leader <- leader[is.finite(leader$age), ]
surv_object <- 
  Surv(
    leader$years, 
    leader$lost,
    type="mstate")
```

```{r subsets-2, fig.width=4.5, fig.height=2.5}
sb3 <- leader$region=="latin america"
km_sb3 <- survfit(surv_object~1, subset=sb3)
ggplot(
  data=tidy(km_sb3), 
  aes(time, estimate, fill=state)) +
  expand_limits(y=1)               +
  geom_col(width=1)
sb4 <- leader$region!="latin america"
km_sb4 <- survfit(surv_object~1, subset=sb4)
ggplot(
  data=tidy(km_sb4), 
  aes(time, estimate, fill=state)) +
  expand_limits(y=1)               +
  geom_col(width=1)
```


```{r save-everything}
save.image("~/survival-models/bin/homework-answers-8.RData")
```